\chapter{Introduction}
\label{ch:bf-introduction}

BayesFilter is a monograph and software-design project for Bayesian estimation
of structural state-space models.  The word structural is important.  The
latent state is not merely an anonymous vector passed to a numerical routine:
its coordinates may be stochastic shocks, deterministic lags, accounting
identities, endogenous completions, auxiliary variables, or measurement-only
summaries.  A filtering backend that ignores those roles can evaluate the
wrong likelihood, produce the wrong gradient, or make a sampler explore states
the model never generates.

The project was extracted from DSGE, macro-finance, and analytic Kalman
derivative work, but its purpose is broader and cleaner than any one source
project.  BayesFilter owns the reusable contracts for state-space likelihoods,
derivatives, diagnostics, and compiled Bayesian sampling.  The economics of NK,
EZ, SGU, CIP, AFNS, and future NAWM-scale applications remain downstream case
studies that stress those contracts.

The organizing premise is that filtering is not a small implementation detail
inside Bayesian state-space inference.  For Hamiltonian Monte Carlo, the filter
is part of the posterior target: it supplies the log likelihood, the gradient
path, the invalid-parameter behavior, and much of the geometry seen by the
sampler.  A filtering backend that is adequate for point likelihood evaluation
may still be unsuitable for HMC if it has hidden non-finite branches, unstable
spectral derivatives, dynamic shapes, structural state violations, or opaque
compilation behavior.

This monograph separates four layers:
\begin{enumerate}[label=(\roman*)]
  \item the structural state-space model contract;
  \item the likelihood and derivative backend;
  \item the HMC target and diagnostic policy;
  \item application case studies that stress the generic contracts.
\end{enumerate}
The separation lets the linear Gaussian and analytic derivative spine be
audited before nonlinear sigma-point filters, particle filters, transport maps,
or DSGE-specific experiments are used as evidence.

\section{Central Design Thesis}
\label{sec:bf-intro-design-thesis}

BayesFilter follows one design thesis:
\[
  \text{model projects own structural maps, and BayesFilter owns filters.}
\]
For an autoregression, the structural map may be a companion-form lag shift.
For a macro-finance model, it may be an affine no-arbitrage state recursion
with masked yields.  For a DSGE model, it may be a perturbation solution that
separates exogenous shocks from endogenous predetermined variables.  These maps
belong to the model project.  The common filtering machinery, derivative
validation, diagnostics, and HMC target discipline belong to BayesFilter.

This separation prevents each application project from rediscovering the same
numerical failure modes.  It also prevents a successful linear Gaussian
implementation from being misused as a nonlinear structural filter simply
because both expose a state vector and covariance matrix.

\section{Provenance and Claim Discipline}
\label{sec:bf-intro-provenance}

The source documents for this project are read-only inputs.  BayesFilter
consolidates them into one notation system and one production-readiness policy,
but it does not treat old prose as automatically authoritative.  Every
substantive claim should be traceable to one of four sources: a local source
chapter or reset memo listed in \texttt{docs/source\_map.yml}, a reviewed
literature artifact, a code/test reference, or an explicitly recorded project
decision such as the TensorFlow Probability NUTS benchmark.

The most important negative rule is equally simple: no backend is described as
HMC-safe merely because it returns a finite likelihood on a smoke test.  HMC
safety requires a target contract, a derivative policy, finite failure handling,
static-shape or compilation discipline where compilation is claimed, and
diagnostics that can reveal when the sampler is exploring a numerically fragile
region.

\section{Reader Map}
\label{sec:bf-reader-map}

The book is written from the stable core outward.  Part~I fixes notation and
interfaces, including structural state partitions.  Part~II develops the exact
linear Gaussian likelihood because it is the regression oracle for more
complicated filters.  Part~III records analytic derivative and custom-gradient
policy before nonlinear filters are promoted into HMC experiments.  Part~IV
discusses nonlinear filters, deterministic completions, and spectral risks with
those derivative requirements already in place.  Part~V states the HMC, JIT,
and diagnostic contracts.  The case studies appear last so they can be judged
against the generic BayesFilter checklist rather than driving the definitions.
