\chapter{State-Space Model Contracts}
\label{ch:bf-state-space-contracts}

BayesFilter treats a state-space model as a contract between a structural model
and a filtering backend.  The structural model may be a DSGE perturbation
solver, an affine term-structure model, a macro-finance system, or a synthetic
validation problem.  The filter should not need to know that origin; it should
receive typed maps with stable shapes and declared derivative policies.

\begin{definition}[BayesFilter state-space contract]
\label{def:bf-ssm-contract}
For parameter vector $\theta \in \Theta \subseteq \R^{d_\theta}$, a
BayesFilter state-space contract supplies latent states $x_t \in \R^{n_x}$,
observations $y_t \in \R^{n_y}$, transition law
\[
  p_\theta(x_{t+1}\mid x_t),
\]
observation law
\[
  p_\theta(y_t\mid x_t),
\]
initial law $p_\theta(x_0)$, and an observation mask or selection rule when the
available components of $y_t$ vary over time.  A linear Gaussian specialization
supplies matrices and vectors
\[
  c_\theta,\quad F_\theta,\quad Q_\theta,\quad
  a_\theta,\quad H_\theta,\quad R_\theta,
\]
with dimensions fixed by the model instance.
\end{definition}

\begin{definition}[Structural state partition]
\label{def:bf-structural-state-partition}
A structural state partition is model metadata that decomposes the latent state
coordinates into declared roles.  The minimal BayesFilter partition records
\[
  x_t = \operatorname{pack}(s_t,d_t,a_t),
\]
where $s_t$ is the stochastic block receiving current innovations directly or
through a declared stochastic transition, $d_t$ is a deterministic-completion
block, and $a_t$ is an optional auxiliary block used for lags, bookkeeping, or
backend-specific summaries.  The partition also records the innovation
dimension and the pack/unpack order.
\end{definition}

The partition is part of the model contract.  It should not be inferred only
from a covariance matrix.  A zero row of a shock-impact matrix is a useful
diagnostic, but it is not a complete timing declaration.  Conversely, a small
positive nugget inserted for numerical regularization does not turn a
deterministic identity into a structural source of randomness.

\begin{assumption}[Shape and value contract]
\label{asm:bf-shape-value-contract}
For a fixed model instance, tensor ranks and leading dimensions are static
throughout a compiled likelihood evaluation.  Parameter-dependent covariance
objects are symmetric positive semidefinite or positive definite as required by
the backend, and invalid parameter values are reported through an explicit
finite-failure policy rather than accidental linear-algebra exceptions.
\end{assumption}

The contract separates model maps from filter maps.  A model map builds
state-space objects from $\theta$.  A filter map consumes those objects and the
data.  A derivative provider supplies first or second derivatives of the
state-space objects, or declares that a backend relies on autodiff through
primitive operations.  This distinction is essential for large models: the
derivative of a structural solver and the derivative of a Kalman or sigma-point
recursion are different mathematical layers.

\begin{remark}[Structural deterministic dynamics are part of the contract]
\label{rem:bf-structural-deterministic-contract}
For DSGE models, the state-space contract must not hide the distinction between
shock-driven exogenous states and endogenous or accounting states that are
deterministic conditional on lagged states and current shocks.  A collapsed
linear transition may be sufficient for an exact Kalman filter, but nonlinear
sigma-point and particle filters need the structural transition map.  This is a
release-gating issue; see
Chapter~\ref{ch:bf-structural-deterministic-dynamics}.
\end{remark}

\section{Degenerate Transitions and Lag Stacks}
\label{sec:bf-degenerate-transitions-lag-stacks}

Structural deterministic coordinates are not special to DSGE models.  An
autoregression of order two already has the same issue.  One companion-form
representation is
\begin{align}
  m_t &= \phi_1 m_{t-1} + \phi_2 \ell_{t-1} + \sigma\varepsilon_t,\\
  \ell_t &= m_{t-1},\\
  x_t &= (m_t,\ell_t).
\end{align}
The lag coordinate $\ell_t$ is deterministic conditional on the previous
state.  In an exact linear Kalman backend, the corresponding transition
covariance is singular:
\[
  Q =
  \begin{bmatrix}
    \sigma^2 & 0\\
    0 & 0
  \end{bmatrix}.
\]
That singularity is structural information, not an implementation failure.
For nonlinear sigma-point or particle filters, quadrature or simulation should
be applied to the innovation or stochastic block and then lifted through the
lag-shift map.

The same contract covers accumulated sums, stock-flow identities,
mixed-frequency aggregation states, and deterministic no-arbitrage recursions.
BayesFilter should therefore test degenerate transitions directly rather than
only through full-rank synthetic examples.

\section{Structural Nonlinear Transition Contract}
\label{sec:bf-structural-nonlinear-transition-contract}

For a nonlinear structural model, a backend should know whether it integrates
over innovations, stochastic states, or the full latent state.  A convenient
contract is
\begin{align}
  s_t &= T_s(s_{t-1},\varepsilon_t;\theta),\\
  d_t &= T_d(d_{t-1},s_{t-1},s_t;\theta),\\
  x_t &= \operatorname{pack}(s_t,d_t,a_t),\\
  y_t &\sim p_\theta(y_t\mid x_t).
\end{align}
When a mixed structural model is evaluated by a full-state Gaussian
approximation, the result must be labeled as an approximation.  The default
nonlinear filter for a mixed structural model should preserve deterministic
identities pointwise for propagated sigma points or particles.

\begin{assumption}[Approximation labeling]
\label{asm:bf-approximation-labeling}
If a backend integrates over a larger stochastic space than the declared model
transition, or injects artificial noise into deterministic coordinates, the
run metadata must record an approximation label.  The label is part of the
likelihood provenance and must be visible to HMC diagnostics and case-study
reports.
\end{assumption}

\section{Masks and Mixed Frequency}
\label{sec:bf-masks-mixed-frequency-contract}

A sparse panel should be represented by a deterministic mask or selection
operator, not by changing the model definition at each time.  At time $t$, the
filter may work with the observed subvector $y_{t,\mathcal{O}_t}$ and the
corresponding selected observation map.  The public contract must state whether
an all-missing time step performs only prediction, whether ragged panels are
batched by pattern, and whether the implementation keeps static shapes for JIT
compilation.

\section{Backend Classes}
\label{sec:bf-backend-classes}

BayesFilter distinguishes the following backend roles:
\begin{itemize}
  \item exact linear Gaussian likelihood backends;
  \item square-root or solve-form variants of exact linear Gaussian backends;
  \item approximate nonlinear Gaussian filters such as EKF, UKF, CKF, and
    SVD-factored sigma-point filters;
  \item particle-filter or simulation-based likelihood estimators;
  \item diagnostic or surrogate backends that do not define the final HMC
    target without an explicit correction policy.
\end{itemize}
The same model can be evaluated by more than one backend, but the monograph
must state which backend defines the posterior target and which backends are
only diagnostic approximations.
