\chapter{BayesFilter API Design}
\label{ch:bf-api-design}

The monograph is not an API reference, but its mathematical contracts should be
precise enough that an implementation can be audited against them.  The API
shape below is therefore a documentation contract rather than a final Python
interface.

\section{Core Objects}
\label{sec:bf-api-core-objects}

A BayesFilter implementation should expose five separable objects:
\begin{itemize}
  \item a parameter transform from sampler coordinates $u$ to model parameters
    $\theta$;
  \item a structural state partition describing stochastic, deterministic, and
    auxiliary coordinates;
  \item a model provider that builds state-space objects from $\theta$;
  \item a filter backend that returns log likelihood and diagnostics;
  \item a derivative provider or autodiff policy that returns the score and,
    when needed, curvature information.
\end{itemize}

For a linear Gaussian model, the model provider returns
\[
  (c,F,Q,a,H,R,m_0,P_0)
\]
and optionally their first and second derivatives with respect to $\theta$.
For a nonlinear model, it returns transition and observation maps together with
the sigma-point, linearization, or simulation policy required by the selected
backend.

\section{Structural Model Protocol}
\label{sec:bf-api-structural-model-protocol}

The exact software interface may change, but implementations should be
auditable against the following documentation contract:
\begin{align*}
  \texttt{partition} &:\quad \text{state names, role indices, innovation dimension},\\
  \texttt{initial\_law} &:\quad (m_0,P_0)\ \text{or a sampling rule},\\
  \texttt{innovation\_law} &:\quad \Sigma_\varepsilon\ \text{or an innovation sampler},\\
  \texttt{unpack\_state},\ \texttt{pack\_state} &:\quad
    x \leftrightarrow (s,d,a),\\
  \texttt{propagate\_stochastic} &:\quad
    (s_{t-1},\varepsilon_t,\theta) \mapsto s_t,\\
  \texttt{complete\_deterministic} &:\quad
    (d_{t-1},s_{t-1},s_t,\theta) \mapsto d_t,\\
  \texttt{observe} &:\quad
    x_t \mapsto p_\theta(y_t\mid x_t).
\end{align*}
Here $s_t$ denotes the declared stochastic block, $d_t$ the deterministic
completion block, and $a_t$ any auxiliary or backend-specific block.  A model
with no deterministic block can set $d_t$ empty; a linear Gaussian companion
form can still expose the partition even if the exact Kalman backend consumes
only the collapsed matrices.

DSGE, MacroFinance, and other application projects should implement this
protocol through adapters.  BayesFilter should not require those projects to
copy filter internals, and those projects should not define private
near-duplicates of BayesFilter filters.

\section{Filter Run Metadata}
\label{sec:bf-api-filter-run-metadata}

Every backend should return metadata that records what target was actually
evaluated.  At minimum, diagnostics should include:
\begin{itemize}
  \item backend name and version;
  \item structural partition summary;
  \item integration space: innovation, stochastic-state, or full-state;
  \item deterministic-completion policy;
  \item approximation label, if any;
  \item derivative policy;
  \item compiled/eager execution status when relevant;
  \item finite-failure or warning codes.
\end{itemize}
This metadata prevents a collapsed approximation, a diagnostic surrogate, or a
partially compiled run from being misreported as a structural HMC target.

\section{Return Contract}
\label{sec:bf-api-return-contract}

The minimum likelihood return is
\[
  (\ell,\ d),
\]
where $\ell$ is a scalar log likelihood and $d$ is a diagnostics record.  A
gradient-capable backend returns
\[
  (\ell,\ \nabla_\theta \ell,\ d),
\]
and a curvature backend returns
\[
  (\ell,\ \nabla_\theta \ell,\ \nabla^2_{\theta\theta}\ell,\ d).
\]
Diagnostics should include enough information to decide whether the result is a
valid target evaluation, a finite invalid-region return, or a numerical warning
that should influence sampler adaptation or backend selection.

\section{Backend Promotion}
\label{sec:bf-api-backend-promotion}

BayesFilter should promote a backend through stages:
\begin{enumerate}[label=(\roman*)]
  \item value-only likelihood evaluation;
  \item gradient validation on small problems;
  \item compiled value-and-gradient execution;
  \item stress tests under masks, boundaries, and near-singular covariances;
  \item HMC diagnostics on controlled targets;
  \item application case studies.
\end{enumerate}
No backend should skip directly from value-only likelihood tests to industrial
HMC use.

\section{Client Adapter Boundary}
\label{sec:bf-api-client-adapter-boundary}

The adapter boundary is intentionally strict.  A client project owns:
\begin{itemize}
  \item structural equations and timing;
  \item parameter transformations and priors;
  \item model-specific observation equations;
  \item perturbation, no-arbitrage, or equilibrium solution objects;
  \item application-level HMC targets and case-study data.
\end{itemize}
BayesFilter owns:
\begin{itemize}
  \item generic filter recursions;
  \item structural partition validation;
  \item square-root and SVD numerical backends;
  \item derivative validation ladders;
  \item finite-failure and diagnostic conventions;
  \item backend promotion gates.
\end{itemize}
The boundary is a maintenance device.  It lets MacroFinance reuse analytic
Kalman derivatives, DSGE reuse structural nonlinear filtering, and future
large models reuse the same validation ladder without forking the numerical
core.
