\chapter{Prediction-Error Decomposition}
\label{ch:bf-prediction-error-decomposition}

The linear Gaussian model is the first BayesFilter reference target because its
likelihood is exact and its derivative recursions can be audited against code
and automatic-differentiation tests.  This chapter records the value-only
likelihood spine; score and Hessian recursions are deferred to
Chapters~\ref{ch:bf-kalman-score}--\ref{ch:bf-kalman-hessian}.

\section{Linear Gaussian Specialization}
\label{sec:bf-lgssm-specialization}

Under the contract in Definition~\ref{def:bf-ssm-contract}, the linear Gaussian
specialization is
\begin{align}
  x_{t+1} &= c(\theta) + F(\theta)x_t + u_{t+1},
  & u_{t+1} &\sim \calN(0,Q(\theta)), \label{eq:bf-lgssm-state} \\
  y_t &= a(\theta) + H(\theta)x_t + \varepsilon_t,
  & \varepsilon_t &\sim \calN(0,R(\theta)). \label{eq:bf-lgssm-observation}
\end{align}
For the exact likelihood backend, $Q(\theta)\succeq 0$ and
$R(\theta)\succ 0$ are required at every accepted target evaluation.  The
initial moments are denoted $m_{0|0}$ and $P_{0|0}$.

The semidefinite allowance for $Q$ is intentional.  Companion-form lag stacks,
mixed-frequency accumulators, and structural deterministic coordinates often
produce process covariances with zero rows or deficient rank.  This is not a
reason to abandon the exact linear Gaussian likelihood.  It is a requirement
that the backend distinguish structural degeneracy from accidental numerical
failure.

\begin{assumption}[Innovation regularity]
\label{asm:bf-innovation-regularity}
For every time point contributing an observation, the selected innovation
covariance $S_t$ defined below is positive definite on the observed
coordinates.  If the model or data mask makes $S_t$ singular, the likelihood
backend must return a named finite-failure diagnostic or use a separately
documented diffuse/singular-observation treatment.
\end{assumption}

\section{Kalman Recursion}
\label{sec:bf-kalman-recursion}

Let
\[
  m_{t|s}=\E[x_t\mid y_{1:s}], \qquad
  P_{t|s}=\Var(x_t\mid y_{1:s}).
\]
The prediction step is
\begin{align}
  m_{t|t-1} &= c + Fm_{t-1|t-1}, \label{eq:bf-kalman-pred-mean}\\
  P_{t|t-1} &= FP_{t-1|t-1}F^\top + Q. \label{eq:bf-kalman-pred-cov}
\end{align}
The innovation and innovation covariance are
\begin{align}
  v_t &= y_t-a-Hm_{t|t-1}, \label{eq:bf-innovation}\\
  S_t &= HP_{t|t-1}H^\top + R. \label{eq:bf-innovation-cov}
\end{align}
When $S_t\succ 0$, the Kalman gain and filtered moments are
\begin{align}
  K_t &= P_{t|t-1}H^\top S_t^{-1}, \label{eq:bf-kalman-gain}\\
  m_{t|t} &= m_{t|t-1}+K_tv_t, \label{eq:bf-kalman-update-mean}\\
  P_{t|t} &= (I-K_tH)P_{t|t-1}. \label{eq:bf-kalman-update-cov}
\end{align}
The covariance update in \eqref{eq:bf-kalman-update-cov} is the compact
textbook form.  Production implementations may use the Joseph form
\[
  P_{t|t}
  =
  (I-K_tH)P_{t|t-1}(I-K_tH)^\top + K_t R K_t^\top
\]
or may update an algebraically equivalent factor rather than materializing a
full covariance matrix.  These are numerical representations of the same
exact linear Gaussian likelihood, not different statistical models.

\section{Log Likelihood}
\label{sec:bf-ped-loglik}

The prediction-error decomposition gives
\begin{equation}
\label{eq:bf-ped-loglik}
  \ell(\theta)
  =
  -\frac{1}{2}\sum_{t=1}^T
  \left[
    \log\det S_t
    + v_t^\top S_t^{-1}v_t
    + n_{y,t}\log(2\pi)
  \right],
\end{equation}
where $n_{y,t}$ is the number of observed components at time $t$.  In a dense
panel, $n_{y,t}=n_y$.  In a masked panel, the same formula applies after
selecting the observed components and the corresponding rows and columns of the
observation objects.

If no components are observed at a time point, the measurement update and
likelihood contribution are skipped and the filter carries the prediction
forward.  This convention treats missingness as part of the observation
contract rather than as a change in the latent transition law.

Equation~\eqref{eq:bf-ped-loglik} is the exact likelihood for
\eqref{eq:bf-lgssm-state}--\eqref{eq:bf-lgssm-observation}.  It is also the
reference value against which approximate nonlinear filters and alternative
linear algebra backends should be tested whenever a controlled linear Gaussian
case is available.

\section{Initialization}
\label{sec:bf-kalman-initialization}

Initial moments are part of the model contract.  If $m_{0|0}$ and $P_{0|0}$ are
fixed by design, derivative initial conditions in later chapters are zero.  If
$P_{0|0}$ is the stationary covariance, then its derivatives are governed by
the derivative Lyapunov equations for the transition system.  BayesFilter must
record which initialization policy defines the target, because changing it
changes the likelihood.

\section{Reference Role}
\label{sec:bf-ped-reference-role}

The prediction-error likelihood is the value oracle for the rest of the
monograph.  Later chapters may evaluate it through covariance, solve-form,
Cholesky, QR, or SVD linear-algebra primitives, but they must agree on
\eqref{eq:bf-ped-loglik} within documented tolerances.  Nonlinear filters
should also recover this value when their model is specialized to a linear
Gaussian case.

Derivative chapters differentiate this same scalar likelihood.  A custom
gradient, analytic score recursion, or Hessian recursion is valid for HMC only
if it corresponds to the value defined here, plus the prior and parameter
transform terms in Chapter~\ref{ch:bf-hmc-target-requirements}.
