\chapter{Stable Linear Filtering}
\label{ch:bf-stable-linear-filtering}

The covariance-form equations in Chapter~\ref{ch:bf-prediction-error-decomposition}
are algebraically correct, but they are not always the best production
primitive.  Large panels, nearly collinear observations, low measurement noise,
and repeated covariance updates can turn small roundoff differences into large
differences in precision matrices and Hessian contractions.

\section{Backend Ladder}
\label{sec:bf-linear-backend-ladder}

BayesFilter uses a backend ladder rather than a single Kalman implementation:
\begin{enumerate}[label=(\roman*)]
  \item covariance-form recursion for clarity and small reference problems;
  \item solve-form likelihood evaluation using factored innovation covariances;
  \item Cholesky square-root propagation where positive pivots are reliable;
  \item QR square-root propagation when reconstruction and solve diagnostics are
    needed;
  \item SVD or spectral fallbacks only with explicit diagnostic labels.
\end{enumerate}
The ladder is a production policy, not a mathematical change to the likelihood.
All exact linear Gaussian backends must agree with
\eqref{eq:bf-ped-loglik} within documented tolerances.

\section{Covariance Form as Reference Algebra}
\label{sec:bf-covariance-form-reference}

The covariance-form recursion is the clearest way to state the likelihood, but
it is not automatically the safest way to compute it.  In finite precision, the
compact covariance update can lose symmetry or positive semidefiniteness after
many time steps.  Implementations should therefore symmetrize only as a
documented numerical projection, not as a way to hide model invalidity.

For small reference tests, covariance form is useful because it exposes the
same moments that appear in the mathematical recursion.  For production
likelihoods, BayesFilter should prefer solve and factor forms that avoid
explicit inverses and provide residual diagnostics.

\section{Solve Form}
\label{sec:bf-solve-form-likelihood}

Let $S_t=L_tL_t^\top$ and define $z_t=L_t^{-1}v_t$.  Then
\[
  \log\det S_t = 2\sum_k \log L_{t,kk},
  \qquad
  v_t^\top S_t^{-1}v_t = z_t^\top z_t.
\]
The likelihood can therefore be evaluated with triangular solves and log
diagonal pivots rather than by explicitly forming $S_t^{-1}$.  This is the
preferred reference form for implementation because the same solves appear in
the analytic score and Hessian recursions.

The solve-form contract is:
\begin{itemize}
  \item factor the selected innovation covariance;
  \item solve for the whitened innovation;
  \item accumulate log determinant and quadratic form from the factor and
    solve result;
  \item record factor pivots and solve residuals.
\end{itemize}
The backend should not form an explicit inverse except in small diagnostic
tests where the numerical error is itself being measured.

\section{Square-Root and Spectral Backends}
\label{sec:bf-square-root-spectral-backends}

Cholesky square-root filters are efficient when positive pivots are reliable.
QR square-root filters are useful when the implementation needs stable stack
updates, reconstruction checks, and a clear path to derivative-factor
validation.  SVD or eigenspectral factorizations are useful for rank-deficient
or nearly rank-deficient linear Gaussian problems, but they should be reported
as spectral backend choices with their own diagnostics.

This distinction matters because the word SVD appears in two different parts
of the monograph.  An SVD-factored exact linear Gaussian backend is still
evaluating \eqref{eq:bf-ped-loglik}.  A nonlinear SVD sigma-point filter is an
approximate nonlinear Gaussian filter whose value and gradient policy require
separate tests.  Success of the former does not certify the latter.

\section{Backend Agreement Tests}
\label{sec:bf-linear-backend-agreement-tests}

Every exact linear Gaussian backend should be tested against at least one
reference backend on controlled models.  A minimal agreement test records:
\begin{itemize}
  \item log-likelihood difference;
  \item maximum filtered-mean difference;
  \item maximum filtered-covariance or reconstructed-factor difference;
  \item minimum pivot or singular value;
  \item solve residual norm;
  \item whether the process covariance is full rank or structurally singular.
\end{itemize}
For singular $Q$ examples, the test should assert that the likelihood is
finite and that the structural degeneracy is reported as model structure or
backend telemetry, not as an unexplained numerical exception.

\section{Diagnostics}
\label{sec:bf-stable-linear-diagnostics}

Stable filtering backends should return diagnostics, not only a scalar
likelihood.  The MacroFinance large-scale backend ladder tests track:
\begin{itemize}
  \item minimum factor pivots;
  \item reconstruction errors for prediction, innovation, and filtered factors;
  \item solve residual norms;
  \item finite log likelihood, gradient, and Hessian values;
  \item backend deltas against the solve-form reference.
\end{itemize}
Those tests classify SVD fallback triggers as no trigger, candidate trigger, or
confirmed trigger.  BayesFilter should preserve that caution: a spectral
fallback is a diagnostic event, not evidence that the HMC gradient path is
automatically safe.

The diagnostic record should travel with the likelihood into HMC reports.  A
finite log likelihood with repeated tiny pivots or frequent fallback triggers
is not the same engineering result as a finite log likelihood with comfortable
factor margins.  Both may be mathematically valid evaluations, but they imply
different sampler risk.
