\chapter{Missing Data and Mixed Frequency}
\label{ch:bf-missing-data-mixed-frequency}

Real macro-finance panels are rarely dense.  Missing observations,
mixed-frequency releases, publication lags, and ragged starts must be handled
as part of the filtering contract rather than as ad hoc data cleaning outside
the likelihood.

\section{Selection Form}
\label{sec:bf-missing-selection-form}

Let $M_t$ be a deterministic selection matrix that extracts the observed
components of $y_t$.  The dense observation equation
\[
  y_t=a+Hx_t+\varepsilon_t,\qquad \varepsilon_t\sim\calN(0,R)
\]
becomes
\[
  M_ty_t = M_ta + M_tHx_t + M_t\varepsilon_t,
  \qquad
  M_t\varepsilon_t\sim\calN(0,M_tRM_t^\top).
\]
The prediction-error likelihood in \eqref{eq:bf-ped-loglik} then uses the
selected innovation and selected innovation covariance.  If no components are
observed at time $t$, the filter performs prediction without a measurement
update.

Equivalently, write $\mathcal{O}_t$ for the observed row set.  The per-period
objects are
\[
  v_{t,\mathcal{O}_t}=y_{t,\mathcal{O}_t}-a_{\mathcal{O}_t}
    -H_{\mathcal{O}_t}m_{t|t-1},
  \qquad
  S_{t,\mathcal{O}_t}
    =H_{\mathcal{O}_t}P_{t|t-1}H_{\mathcal{O}_t}^\top
      +R_{\mathcal{O}_t,\mathcal{O}_t}.
\]
The likelihood contribution uses the dimension
$|\mathcal{O}_t|$.  If $\mathcal{O}_t$ is empty, the contribution is zero and
the filtered moments equal the predicted moments.  This convention is important
for ragged panels because it makes missingness explicit in the likelihood
rather than hidden in preprocessed data.

\section{Mixed-Frequency State Augmentation}
\label{sec:bf-mixed-frequency-state-augmentation}

Mixed-frequency models often require deterministic auxiliary states.  Examples
include cumulators for quarterly growth, publication-lag states, and running
averages used to align daily, monthly, and quarterly releases.  These
coordinates are part of the structural state contract from
Definition~\ref{def:bf-structural-state-partition}.  Their process covariance
may be singular because the auxiliary rows are deterministic functions of
lagged states and current high-frequency states.

BayesFilter should therefore distinguish two operations:
\begin{enumerate}[label=(\roman*)]
  \item augmenting the state so the observation equation is linear or
    conditionally linear at the target sampling frequency;
  \item selecting the observed rows at each time point.
\end{enumerate}
Both operations preserve the prediction-error likelihood when the augmented
transition and observation maps are the declared model.  A small ridge added
only for numerical conditioning must be recorded as numerical regularization,
not as a new source of structural shock variation.

\section{Implementation Policy}
\label{sec:bf-missing-implementation-policy}

The public backend must state whether it supports sparse panels.  A dense-only
backend should fail before likelihood evaluation if the observations contain
NaNs, infinities, or masks that imply unsupported missing-data behavior.  The
MacroFinance missing-data policy tests explicitly check that unsupported sparse
observation masks raise a named pending-support error rather than leaking NaNs
into the Kalman recursion.

For compiled likelihoods, there are two viable policies:
\begin{itemize}
  \item group time steps by mask pattern and keep each compiled path static;
  \item keep a fixed dense shape and use selection or masking operations whose
    shape is known at trace time.
\end{itemize}
Either policy is acceptable if the likelihood value and derivative contract are
unchanged and the diagnostics identify which observations contributed to each
time step.

The diagnostic record should include at least:
\begin{itemize}
  \item number of observed rows at each time or by mask pattern;
  \item count of all-missing prediction-only steps;
  \item mask-pattern identifiers used for compiled grouping;
  \item whether missingness was handled by row selection, dense masking, or an
    unsupported fallback;
  \item any regularization applied to augmented deterministic rows.
\end{itemize}

\section{Derivative Deferral}
\label{sec:bf-missing-derivative-deferral}

The value contract above is independent of the derivative implementation.
Later derivative chapters must state whether mask patterns are differentiated
as fixed data, whether all-missing steps contribute zero derivative to the
measurement likelihood, and whether mixed-frequency augmentation parameters
enter the transition, the observation map, or both.  This chapter fixes the
likelihood object; it does not yet certify any derivative backend.
