\chapter{Large-Scale Linear Gaussian State-Space Models}
\label{ch:bf-large-scale-lgssm}

Large linear Gaussian models are the bridge between textbook Kalman filtering
and industrial Bayesian state-space inference.  They are still exact, but their
dimension exposes the numerical and compilation issues that small examples can
hide.

\section{Scale Targets}
\label{sec:bf-lgssm-scale-targets}

A BayesFilter scale target should record:
\begin{itemize}
  \item state dimension $n_x$;
  \item observation dimension $n_y$;
  \item time length $T$;
  \item parameter dimension $d_\theta$;
  \item dense or sparse observation policy;
  \item backend ladder and fallback rules;
  \item memory layout and compilation policy.
\end{itemize}
The MacroFinance large-scale LGSSM tests include scenarios such as
\texttt{baseline\_10x3x5}, low-measurement-noise collinearity,
ill-conditioned process noise, and near-rank-loss candidates.  BayesFilter
uses these as infrastructure evidence, not as a claim that every NAWM-sized
model is already solved.

\section{Backend Registry}
\label{sec:bf-lgssm-backend-registry}

Large-scale runs should be described by a backend registry rather than by a
single method name.  A registry entry should identify:
\begin{itemize}
  \item the value backend, such as covariance form, solve form, Cholesky
    square root, QR square root, or spectral fallback;
  \item the derivative backend, such as tape autodiff, analytic score,
    analytic Hessian, custom gradient, or finite-difference diagnostic;
  \item the initialization policy;
  \item the missing-data policy;
  \item static-shape and compilation assumptions;
  \item finite-failure diagnostics and fallback labels.
\end{itemize}
This registry is part of the scientific result.  A reported HMC run without
backend metadata is not reproducible enough to diagnose whether a divergence,
small effective sample size, or slow compilation came from the posterior
geometry, the filtering algebra, the derivative path, or the execution mode.

\section{Validation Ladder}
\label{sec:bf-lgssm-validation-ladder}

The validation ladder should move from exact values to sampler behavior:
\begin{enumerate}[label=(\roman*)]
  \item value parity against the prediction-error decomposition in
    Chapter~\ref{ch:bf-prediction-error-decomposition};
  \item filtered-moment and backend-delta parity across covariance, solve, and
    square-root implementations on controlled small cases;
  \item stress cases with low measurement noise, nearly collinear observations,
    singular or ill-conditioned process covariance, and long panels;
  \item derivative parity on small cases before the derivative backend is used
    by HMC;
  \item compiled-path checks for static shape, finite diagnostics, and stable
    recompilation behavior;
  \item short HMC smoke runs that exercise the target without being used as
    convergence evidence;
  \item medium and strict HMC recovery only after the value and derivative
    gates have passed.
\end{enumerate}
The same ladder should be reused for MacroFinance adapters, DSGE adapters, and
future BayesFilter-native examples.  Passing it for one model family is useful
evidence about the backend, but it does not automatically transfer to another
state partition, observation mask policy, or parameterization.

\section{Stress Scenarios}
\label{sec:bf-lgssm-stress-scenarios}

The stress suite should include both numerical and modeling edge cases:
\begin{itemize}
  \item structurally singular $Q$ from AR($p$) companion states,
    mixed-frequency accumulators, or deterministic structural coordinates;
  \item nearly singular innovation covariance from collinear measurements or
    very small measurement noise;
  \item long panels where repeated updates test covariance symmetry and factor
    reconstruction;
  \item ragged panels and all-missing time steps;
  \item parameter draws near admissibility boundaries;
  \item backend fallback triggers and spectral-gap telemetry.
\end{itemize}
Each scenario should define the expected diagnostic outcome before it is used
as an HMC target.  Some cases should pass with comfortable numerical margins;
others should return named finite failures.  Both outcomes are useful if they
are intentional and reproducible.

\section{Readiness Criteria}
\label{sec:bf-lgssm-readiness}

A large-scale LGSSM backend is ready for HMC experiments only after:
\begin{enumerate}[label=(\roman*)]
  \item the value backend matches a solve-form or covariance-form oracle;
  \item gradient and Hessian backends match finite differences or autodiff on
    small controlled cases;
  \item square-root or QR backends match the solve-form backend within stated
    tolerances on dimension-scaled cases;
  \item diagnostics remain finite under stress scenarios;
  \item shape and compilation behavior are documented.
\end{enumerate}

The current evidence supports this ladder as a validation design.  It does not
remove the need to rerun the ladder for each new model family, parameterization,
or missing-data policy.

\section{Derivative Consolidation Hypotheses}
\label{sec:bf-lgssm-derivative-hypotheses}

The next analytic-derivative pass should test the following hypotheses against
the MacroFinance derivation and implementation:
\begin{enumerate}[label=(\roman*)]
  \item the Chapter~\ref{ch:bf-prediction-error-decomposition} value contract
    is identical to the likelihood differentiated in the MacroFinance analytic
    Kalman note;
  \item analytic score recursions match autodiff or finite differences on small
    dense and singular-$Q$ cases;
  \item Hessian recursions match finite-difference-on-gradient diagnostics on
    small cases before they are used for mass-matrix or curvature proposals;
  \item solve-form derivatives avoid explicit innovation-covariance inverses in
    production code;
  \item derivative diagnostics identify the first time step and matrix block
    responsible for nonfinite values or large parity errors.
\end{enumerate}
These are hypotheses, not completed certification.  They should be promoted to
claims only after source derivations, code, and numerical tests agree.
