\chapter{Extended Kalman Filtering}
\label{ch:bf-ekf}

The extended Kalman filter is the most direct nonlinear extension of the
linear Gaussian spine in Chapter~\ref{ch:bf-prediction-error-decomposition}.
It replaces nonlinear transition and observation maps by local linear
approximations and then applies the same prediction-error likelihood form to
the resulting Gaussian approximation.  In BayesFilter this is an approximate
likelihood backend, not an exact replacement for the Kalman filter except in
the affine-Gaussian special case.

\section{Local Linearization}
\label{sec:bf-ekf-local-linearization}

Consider the nonlinear contract
\begin{align}
  x_{t+1} &= f_\theta(x_t) + u_{t+1},
  & u_{t+1} &\sim \calN(0,Q_\theta), \label{eq:bf-ekf-transition}\\
  y_t &= h_\theta(x_t) + \varepsilon_t,
  & \varepsilon_t &\sim \calN(0,R_\theta). \label{eq:bf-ekf-observation}
\end{align}
Let
\[
  F_t =
  \left.\frac{\partial f_\theta(x)}{\partial x^\top}
  \right|_{x=m_{t-1|t-1}},
  \qquad
  H_t =
  \left.\frac{\partial h_\theta(x)}{\partial x^\top}
  \right|_{x=m_{t|t-1}} .
\]
The EKF prediction and update are
\begin{align}
  m_{t|t-1} &= f_\theta(m_{t-1|t-1}), \\
  P_{t|t-1} &= F_tP_{t-1|t-1}F_t^\top+Q_\theta, \\
  v_t &= y_t-h_\theta(m_{t|t-1}), \\
  S_t &= H_tP_{t|t-1}H_t^\top+R_\theta, \\
  K_t &= P_{t|t-1}H_t^\top S_t^{-1}, \\
  m_{t|t} &= m_{t|t-1}+K_tv_t .
\end{align}
The covariance may be updated in simplified form or in a Joseph-style
linearized form.  The backend must state which form defines the numerical
recursion, because this choice affects finite-arithmetic behavior even when the
first-order algebra is equivalent.

\section{Likelihood Status}
\label{sec:bf-ekf-likelihood-status}

The EKF log likelihood uses the same innovation expression as
\eqref{eq:bf-ped-loglik}, but the innovations and innovation covariances come
from local linearization.  Therefore the EKF target is an approximate target
unless $f_\theta$ and $h_\theta$ are affine on the region reached by the
filter.  An HMC chain using the EKF samples from the posterior induced by that
approximate likelihood and prior, not from the exact nonlinear state-space
posterior.

For BayesFilter documentation and tests, the EKF should be used in three roles:
\begin{enumerate}[label=(\roman*)]
  \item a baseline nonlinear Gaussian approximation;
  \item a regression oracle in cases where the nonlinear maps collapse to
    affine maps and the exact Kalman likelihood is available;
  \item a diagnostic comparison for sigma-point and particle backends.
\end{enumerate}

\section{HMC Contract}
\label{sec:bf-ekf-hmc-contract}

An EKF backend is HMC-eligible only after it declares:
\begin{itemize}
  \item how the Jacobians $F_t$ and $H_t$ are computed;
  \item whether gradients with respect to $\theta$ are tape-based, analytic, or
    custom;
  \item what happens when $S_t$ is singular, non-finite, or numerically
    indefinite;
  \item whether all loop lengths, masks, and state dimensions remain static
    under compilation;
  \item which parity tests compare the EKF to exact linear Gaussian cases.
\end{itemize}
The older nonlinear-filtering chapters in the source projects motivate this
contract, but BayesFilter tightens it for HMC: differentiability and finite
failure behavior are part of the target definition, not implementation
afterthoughts.
