\chapter{Square-Root Sigma-Point Filters}
\label{ch:bf-square-root-sigma-point}

Square-root sigma-point filters propagate a factor of the covariance rather
than the covariance matrix itself.  Their purpose is numerical: avoid avoidable
loss of symmetry and positive semi-definiteness, reduce solve instability, and
make the filtering recursion easier to monitor.  In BayesFilter, square-root
bookkeeping is a backend discipline.  It does not by itself make an
approximate likelihood exact or an autodiff gradient safe for HMC.

\section{Backend Contract}
\label{sec:bf-square-root-contract}

A square-root backend stores a factor $C_t$ such that
$P_t=C_tC_t^\top$ or an equivalent factorization.  The factor may be Cholesky,
QR-derived, SVD/eigen-derived, or another auditable representation.  The
backend contract must state:
\begin{itemize}
  \item the factor orientation and reconstruction identity;
  \item the predict update and measurement update used for the factor;
  \item whether covariance matrices are ever reconstructed;
  \item how rank loss, jitter, and PSD projection are handled;
  \item which operations define the differentiable target path.
\end{itemize}

The source SR-UKF material is useful here precisely because it exposed a
maintenance risk: a method can be called square-root while still reconstructing
full covariances and applying simplified covariance updates internally.  The
BayesFilter monograph should therefore describe the actual implemented
recursion, not only the label attached to the filter.

\section{HMC-Relevant Risks}
\label{sec:bf-square-root-hmc-risks}

For HMC, the relevant question is not only whether the factor recursion keeps
$P_t$ positive semi-definite in floating point.  The sampler also needs a
stable derivative of the scalar log target.  Cholesky, QR, and SVD factors have
different derivative behavior near rank loss or repeated singular values.  A
backend that is robust for value evaluation can still be fragile as a raw
tape-gradient path.

BayesFilter should therefore classify square-root sigma-point evidence in two
columns:
\begin{enumerate}[label=(\roman*)]
  \item value-recursion evidence, such as finite likelihoods, factor
    reconstruction residuals, and affine-Gaussian parity;
  \item derivative evidence, such as finite gradients, AD/custom-gradient
    parity, spectral-gap telemetry, and HMC leapfrog stability.
\end{enumerate}
Only the second column is directly relevant to HMC production readiness.

\section{When to Prefer Square-Root Forms}
\label{sec:bf-square-root-choice}

Square-root forms are natural when observation noise is small, latent
covariances are nearly singular, or long scans accumulate asymmetry.  They are
also useful for large LGSSMs where factor diagnostics expose numerical problems
earlier than a final log likelihood does.  They should not be treated as a
universal answer to nonlinear filtering: if the moment closure is poor, the
filter is still approximating the wrong target; if the derivative of the
factorization is unstable, HMC can still fail.
