\chapter{SVD Sigma-Point Filters}
\label{ch:bf-svd-sigma-point}

The SVD sigma-point filter is the most important nonlinear-filtering lesson
from the recent source-project debugging work.  It improved value-side
robustness by using spectral covariance factors and explicit finite-arithmetic
guards, but it also exposed why raw autodiff through spectral decompositions is
not a comfortable industrial HMC default.  BayesFilter should preserve both
lessons.

\section{What the Source Audit Established}
\label{sec:bf-svd-audit-established}

The math/code audit recorded in
the source-project SVD filter audit result from 2026-04-25
established the following bounded facts:
\begin{itemize}
  \item the linear SVD Kalman path uses a Joseph-style covariance update;
  \item the nonlinear sigma-point path uses the simplified update
    \[
      P^+ = P^- - K P_{zz} K^\top ;
    \]
  \item PSD projection occurs after the candidate covariance is formed, so it
    cannot prevent overflow or non-finite values during the formation of
    $P^+$;
  \item the conditioning guard with threshold
    \texttt{\_SVD\_CONDITIONING\_MAX\_ABS = 1e100} is an engineering
    finite-arithmetic guard, not a mathematically derived posterior threshold;
  \item prior short HMC diagnostics were numerical-path regressions, not
    convergence evidence.
\end{itemize}
These are implementation facts, not general theorems about SVD filtering.
BayesFilter should cite them as project evidence and should avoid converting
them into unsupported production-readiness claims.

\section{Spectral Differentiation Risk}
\label{sec:bf-svd-spectral-risk}

The Phase 2B literature gate accepted the Matrix Backprop source for the
bounded claim that gradients through SVD or eigendecomposition contain terms
with denominators involving singular-value or eigenvalue gaps.  When two
singular values or eigenvalues are equal, the decomposition is not uniquely
differentiable; when they are merely close, derivative terms can become large
and numerically unstable.

This is directly relevant to HMC.  A large DSGE or NAWM-sized model can visit
parameter regions where covariance factors are ill conditioned, nearly rank
deficient, or spectrally clustered.  A value-side SVD factor may still return a
finite covariance approximation, while the tape gradient through the spectral
basis becomes noisy, discontinuous, or non-finite.  Larger dimensions make
close spectral gaps more likely in practice, so BayesFilter should require
gap telemetry and derivative validation before using SVD sigma-point gradients
inside production HMC.

\section{Status Labels for SVD-HMC}
\label{sec:bf-svd-status-labels}

The SVD/HMC gap-closure plan introduced four labels that BayesFilter should
reuse:
\begin{description}
  \item[filter-correct] the likelihood and gradients agree with a reference
    behavior on controlled cases;
  \item[sampler-usable] short or medium HMC runs produce finite chains and
    useful diagnostics;
  \item[converged] strict multi-chain posterior diagnostics pass on the
    intended target;
  \item[production-XLA-gated] fixed-solution, full-solve, boundary rejection,
    gradient, and tiny HMC paths compile and run under the production
    compilation policy.
\end{description}
These labels prevent a recurring mistake: a clean smoke run is not a
convergence result, and a finite likelihood is not a validated HMC gradient.

\section{BayesFilter Policy}
\label{sec:bf-svd-policy}

BayesFilter may use SVD sigma-point filters as robust approximate likelihood
infrastructure and as a diagnostic backend.  To use them as HMC production
targets, the project must additionally provide:
\begin{enumerate}[label=(\roman*)]
  \item affine-Gaussian recovery against the exact Kalman backend;
  \item nonlinear controlled recovery against dense quadrature or a trusted
    reference;
  \item finite value and finite gradient stress tests across invalid and
    near-boundary parameter proposals;
  \item eigenvalue or singular-value gap telemetry during HMC;
  \item eager/compiled parity for the complete value-and-gradient path;
  \item a custom-gradient or analytic-gradient policy if raw spectral autodiff
    fails the preceding checks.
\end{enumerate}
For industrial-size models, the conservative default is prevention rather than
debugging: certify a derivative path before relying on SVD-tape gradients in a
long HMC run.
