\chapter[Structural DSGE Dynamics]{Structural Deterministic Dynamics in DSGE Filtering}
\label{ch:bf-structural-deterministic-dynamics}

\begin{mdframed}
\textbf{Production warning.}
Do not implement DSGE nonlinear filtering by silently treating every declared
state coordinate as an exchangeable noisy Gaussian coordinate.  Many DSGE
models contain a structural split between shock-driven exogenous states and
endogenous or accounting states that are deterministic conditional on lagged
states, controls, and current shocks.  Collapsing that structure into a generic
additive transition can be harmless for a linear Kalman filter, but it is a
structural design error for sigma-point and particle filters unless an explicit
approximation argument is written down and tested.
\end{mdframed}

\section{The Structural Split}
\label{sec:bf-dsge-structural-split}

The state vector in a DSGE likelihood is not merely a numerical vector.  It
usually carries economic timing.  In the language of the structural state
partition from Chapter~\ref{ch:bf-state-space-contracts}, a common DSGE
representation separates a declared stochastic block from a deterministic-
completion block.  Writing the stochastic block as exogenous state variables
$m_t$ and the deterministic-completion block as endogenous predetermined or
accounting state variables $k_t$, one common representation is:
\begin{align}
  m_t &= T_m(m_{t-1}, \varepsilon_t; \theta), \label{eq:bf-m-transition}\\
  k_t &= T_k(k_{t-1}, m_{t-1}, m_t; \theta), \label{eq:bf-k-transition}\\
  x_t &= (m_t, k_t).
\end{align}
In a perturbation solution, this structure may also be represented by a compact
law of motion
\[
  x_t = h_x x_{t-1} + \eta \varepsilon_t
\]
at first order, with zero or rank-deficient innovation rows for the
deterministic-completion block.  The compact representation is useful, but it
must not erase the timing contract or the structural metadata described in
Chapter~\ref{ch:bf-api-design}.  For mixed structural models, the one-step
predictive law is the pushforward of the joint law of the lagged state and the
current stochastic input through the structural transition map:
\[
  p_\theta(x_t\mid y_{1:t-1})
  =
  \iint \delta\!\left(x_t - F_\theta(x_{t-1},\varepsilon_t)\right)
  p(\varepsilon_t)\,p_\theta(x_{t-1}\mid y_{1:t-1})
  \, d\varepsilon_t \, dx_{t-1}.
\]
In nonlinear filtering, the correct random input is therefore the innovation or
declared stochastic block.  The endogenous block is then generated by the
model-implied deterministic-completion map.  Here ``deterministic'' means no
independent innovation in that coordinate once $(x_{t-1},\varepsilon_t)$ is
fixed; it does not mean zero predictive variance conditional only on the lagged
filtering state.

This is the distinction emphasized in the DSGE filtering literature.  It is
also an instance of the generic BayesFilter structural-transition contract from
Chapter~\ref{ch:bf-state-space-contracts}: the backend integrates over the
declared stochastic block and completes the deterministic block pointwise.
Standard Bayesian DSGE expositions write the likelihood in state-space form
after the model solution has supplied a transition law for predetermined
variables and an observation equation for measured macroeconomic series
\citep{herbst2015bayesian,an2007bayesian}.  Pruned perturbation methods make
the same point in a different language: the first-order component is propagated
stochastically, while the higher-order correction is propagated by a separate
deterministic recursion driven by the lower-order state
\citep{kim2008calculating,andreasen2018pruned}.  The filtering backend must
honor the state-space law supplied by the structural solution, not merely the
dimension of the state vector.

\section{Why Linear Kalman Filtering Can Hide the Problem}
\label{sec:bf-linear-kalman-hides-structural-split}

For a linear Gaussian state-space model,
\begin{align}
  x_t &= c + F x_{t-1} + u_t,\qquad u_t \sim \calN(0,Q),\\
  y_t &= a + H x_t + e_t,\qquad e_t \sim \calN(0,R),
\end{align}
the Kalman filter only needs the conditional mean and covariance of
$x_t\mid x_{t-1}$.  If a DSGE solution implies a singular or nearly singular
$Q$, the exact linear Gaussian likelihood remains well defined when the
collapsed representation preserves the same conditional law induced by the
structural transition.  Numerically stable implementations may then handle that
degenerate conditional Gaussian law through diffuse, square-root, solve-form,
or carefully regularized recursions as appropriate \citep{durbin2012time}.

This is why the following collapsed first-order representation can be
acceptable in an exact Kalman path:
\[
  Q = \eta \eta^\top,\qquad
  P_{t|t-1} = h_x P_{t-1|t-1} h_x^\top + Q.
\]
The key point is not that structural determinism disappears.  The key point is
that the collapsed linear representation still encodes the same one-step
conditional Gaussian law.  Deterministic-completion coordinates are not wrong
merely because they are included in $x_t$.  They obtain randomness through
their dependence on lagged stochastic coordinates, and their direct innovation
variance may be zero.  The exact linear Kalman recursion uses only the correct
first two conditional moments.

The danger is that this success can train the implementation culture into a
bad habit: it makes the state vector look like a homogeneous Gaussian vector.
That habit becomes wrong when the backend changes from exact linear filtering
to nonlinear sigma-point or particle filtering.  The relevant design decision
is then the integration space recorded in the filter metadata, not merely the
numerical dimension of $x_t$.

A related misunderstanding should be removed explicitly.  A deterministic-
completion coordinate need not have zero one-step predictive variance given the
lagged state.  It can inherit randomness through the current stochastic block.
In the Chapter~\ref{ch:bf-state-space-contracts} AR(2) lag-stack example, the
lag coordinate has no new innovation of its own but still remains random under
the lagged filtering distribution.  Likewise, in DSGE models with
\eqref{eq:bf-m-transition}--\eqref{eq:bf-k-transition}, $k_t$ may satisfy
\[
  \Var(k_t\mid x_{t-1}) > 0
\]
even though $k_t$ has no independent innovation once $(x_{t-1},\varepsilon_t)$
is fixed.

\section{Why Nonlinear Filters Must Respect the Structural Map}
\label{sec:bf-nonlinear-filters-need-structural-map}

Sigma-point and particle filters approximate a transition law, not just a
matrix covariance update.  In a generic nonlinear state-space model,
\[
  x_t = f_\theta(x_{t-1}, \varepsilon_t),\qquad
  y_t = g_\theta(x_t, e_t),
\]
the filter has to decide which variables are integrated over and which are
deterministic transforms.  In BayesFilter language, this is the choice of
integration space recorded in the filter metadata.  For DSGE models with
\eqref{eq:bf-m-transition}--\eqref{eq:bf-k-transition}, the natural nonlinear
prediction step is:
\begin{enumerate}[label=(\roman*)]
  \item represent the filtering distribution of the lagged structural state;
  \item draw, integrate, or choose quadrature points for the exogenous
    innovation $\varepsilon_t$ or current exogenous state $m_t$;
  \item compute each corresponding endogenous state $k_t$ using
    $T_k(k_{t-1},m_{t-1},m_t;\theta)$;
  \item evaluate the observation map on the resulting structurally valid
    points.
\end{enumerate}
The endogenous coordinates should not receive arbitrary independent
``sigma-point noise'' merely because they are entries of $x_t$.  If a sigma
point moves an accounting identity or predetermined endogenous state outside
the model-implied support of the structural transition, the filter is no longer
approximating the intended DSGE state-space model.  It is approximating a
different artificial model.  This is exactly the approximation boundary from
Chapter~\ref{ch:bf-sigma-point-filters}: quadrature may approximate the target
law, but injecting additional stochastic coordinates changes the target unless
that change is declared and justified.

Particle-filter literature makes the same distinction operationally.  A
particle filter propagates particles through the state transition and weights
them by the observation density \citep{gordon1993novel,doucet2001sequential,
andrieu2010particle}.  Written in the structural notation, a basic propagation
and weighting step is
\[
  x_t^{(i)} = F_\theta\!\left(x_{t-1}^{(a_i)},\varepsilon_t^{(i)}\right),
  \qquad
  w_t^{(i)} \propto p_\theta\!\left(y_t\mid x_t^{(i)}\right).
\]
When part of the transition is deterministic, particles should be propagated by
sampling the declared stochastic block and then deterministically completing the
remaining coordinates.  Adding artificial noise to deterministic coordinates can
be used only as a declared proposal or regularization device, with the
proposal/target correction and approximation status stated explicitly.  It is
not the default filtering law.

\section{Reusable BayesFilter Structural Filtering Step}
\label{sec:bf-structural-filter-step-algorithm}

The doctrine above can be stated as a reusable backend contract.  At every time
step, the filter should treat the structural transition as the object being
approximated:
\[
  x_t = F_\theta(x_{t-1},\varepsilon_t),
  \qquad
  y_t \sim p_\theta(y_t\mid x_t).
\]
The filtered state is still the full state $x_t$.  The integration variable is
the pre-transition uncertainty that generates the predictive law.

\begin{mdframed}
\textbf{Algorithm: \texttt{structural\_filter\_step}.}

\begin{enumerate}[label=(\arabic*)]
  \item Start from the filtered approximation to
    $p_\theta(x_{t-1}\mid y_{1:t-1})$ and the declared innovation law
    $p_\theta(\varepsilon_t)$.
  \item Form sigma points, quadrature points, particles, or linearization
    points in the declared integration space:
    $(x_{t-1},\varepsilon_t)$, $(s_{t-1},d_{t-1},\varepsilon_t)$, or an
    equivalent stochastic-state parameterization.
  \item For each point, call the structural transition map:
    \[
      s_t^{(j)}=T_s(s_{t-1}^{(j)},\varepsilon_t^{(j)};\theta),
      \qquad
      d_t^{(j)}=T_d(d_{t-1}^{(j)},s_{t-1}^{(j)},s_t^{(j)};\theta).
    \]
    Deterministic-completion coordinates are outputs of this map, not
    independent new stochastic coordinates.
  \item Pack $x_t^{(j)}=\operatorname{pack}(s_t^{(j)},d_t^{(j)},a_t^{(j)})$
    and evaluate the observation map or density on these structurally valid
    points.
  \item Apply the selected update: exact Kalman, Gaussian sigma-point closure,
    particle weighting/resampling, or another declared backend.
  \item Return the likelihood contribution, filtered approximation, and
    metadata recording integration space, deterministic-completion policy,
    approximation label, derivative policy, and finite-failure diagnostics.
\end{enumerate}
\end{mdframed}

This algorithm is intentionally backend-neutral.  For a linear Gaussian Kalman
backend, the structural transition may collapse to matrices $(F,Q)$ as long as
the same conditional Gaussian law is preserved.  For UKF, CKF, SVD sigma-point,
and particle backends, the pointwise structural transition is the default.  A
mixed-model full-state Gaussian approximation may still be offered as a
diagnostic or legacy path, but it must carry an approximation label and must be
tested against structural references.

\section[Degenerate Transitions]{Degenerate Transitions Are Structural}
\label{sec:bf-degenerate-transitions-structural}

Degenerate transition covariance is not an exceptional corner case in DSGE
models.  It is often the mathematical expression of timing, identities, and
endogenous accumulation.  Capital, debt, lagged output growth, risk-premium
terms, and other predetermined variables may be deterministic conditional on
the previous state and current shocks.  A full covariance matrix with small positive diagonal padding can keep linear
algebra routines alive, but it does not by itself define the intended economic
transition.  BayesFilter therefore separates three ideas that are often
confused in practice:
\begin{enumerate}[label=(\roman*)]
  \item an exact degenerate transition law that already matches the structural
    model;
  \item numerical regularization used to evaluate that law stably; and
  \item an altered transition law that injects new stochastic degrees of
    freedom into deterministic-completion coordinates.
\end{enumerate}
Only the first two preserve the original structural model automatically.

The implementation policy is therefore:
\begin{itemize}
  \item distinguish structural determinism from numerical regularization;
  \item never let a nugget or PSD projection change the declared stochastic
    dimension without a written approximation label;
  \item expose deterministic blocks in diagnostics, rather than burying them
    under a full-rank covariance;
  \item test degenerate and rank-deficient transition cases directly.
\end{itemize}

This point is separate from the SVD spectral-gradient issue in
Chapter~\ref{ch:bf-svd-sigma-point}.  SVD factors may improve value-side
robustness for nearly singular covariances.  They do not repair a filter that
is propagating the wrong state-space law.

\section{Worked Structural UKF Example}
\label{sec:bf-structural-ukf-worked-example}

This section works through a one-step unscented update in a model with one
shock-driven state and one deterministic endogenous state.  The example is
small enough to expose the algebra, but it contains the same structural issue
as a DSGE model with exogenous shocks and endogenous predetermined variables.
The point of the example is not to prove a universal theorem about every UKF
variant.  The point is to show the BayesFilter structural-design rule for mixed
structural models: generate sigma points in the declared pre-transition
uncertainty variables, then complete deterministic coordinates pointwise.

Before turning to the arithmetic, it is useful to isolate what differs from a
generic additive-noise full-state UKF pattern.  In many textbook UKF
implementations, one starts from a Gaussian approximation for the full state and
a process-noise covariance, generates sigma points for that enlarged Gaussian
object, and then propagates every state coordinate as if it belonged to one
homogeneous stochastic transition.  That pattern can be a usable approximation
in some models, but for mixed structural models it is not automatically the same
target as the structural transition declared in
Chapter~\ref{ch:bf-state-space-contracts}.

For the structural DSGE transition
\eqref{eq:bf-m-transition}--\eqref{eq:bf-k-transition}, the contrast is:
\begin{itemize}
  \item \textbf{Structural UKF:} generate sigma points in the declared
    pre-transition uncertainty variables, typically $(x_{t-1},\varepsilon_t)$
    or the declared stochastic block together with the lagged state, then
    complete the deterministic-completion block pointwise.
  \item \textbf{Naive full-state UKF approximation:} generate sigma points for
    a full post-transition state covariance that grants every coordinate a
    direct stochastic perturbation, even when the structural model declares some
    coordinates to belong to the deterministic-completion block.
\end{itemize}
The second construction is not automatically useless, but in this class of
models it enlarges the stochastic space.  It therefore requires an explicit
approximation label and a justification against a reference if it is used at
all.

\subsection{Why the Sigma-Point Variable Uses Pre-Transition Uncertainty}
\label{sec:bf-structural-ukf-proposition}

\begin{proposition}[Predictive pushforward and sigma-point space]
\label{prop:bf-structural-ukf-pushforward}
Under the structural transition
\begin{align}
  m_t &= T_m(m_{t-1},\varepsilon_t),
    \label{eq:bf-structural-ukf-prop-m}\\
  k_t &= T_k(k_{t-1},m_{t-1},m_t),
    \label{eq:bf-structural-ukf-prop-k}\\
  x_t &= (m_t,k_t),
    \label{eq:bf-structural-ukf-prop-state}
\end{align}
with innovation law $p(\varepsilon_t)$ independent of
$x_{t-1}\mid y_{1:t-1}$, the following hold:
\begin{enumerate}[label=(\roman*)]
  \item the one-step predictive law $p(x_t\mid y_{1:t-1})$ is the pushforward
    of the joint law of $(x_{t-1},\varepsilon_t)$ under the structural map;
  \item a UKF that approximates this predictive law should place sigma points on
    $(x_{t-1},\varepsilon_t)$, or on an equivalent parameterization of the same
    pre-transition uncertainty;
  \item a naive full-state additive-noise UKF is exact only if its induced
    predictive law coincides with that pushforward law; if it adds a direct
    perturbation to $k_t$, it targets a different law.
\end{enumerate}
\end{proposition}

\begin{proof}
Define the structural map
\begin{equation}
  F(x_{t-1},\varepsilon_t)
  =
  \bigl(T_m(m_{t-1},\varepsilon_t),
  T_k(k_{t-1},m_{t-1},T_m(m_{t-1},\varepsilon_t))\bigr).
  \label{eq:bf-structural-ukf-prop-map}
\end{equation}
Because the current innovation is independent of the lagged filtering law, the
joint pre-transition uncertainty factorizes as
\begin{equation}
  p(x_{t-1},\varepsilon_t\mid y_{1:t-1})
  =
  p(x_{t-1}\mid y_{1:t-1})p(\varepsilon_t).
  \label{eq:bf-structural-ukf-prop-factorization}
\end{equation}
Let $\varphi$ be any bounded test function on the state space of $x_t$.  Then
\begin{equation}
  \E\!\left[\varphi(x_t)\mid y_{1:t-1}\right]
  =
  \iint
  \varphi\!\left(F(x_{t-1},\varepsilon_t)\right)
  p(x_{t-1}\mid y_{1:t-1})p(\varepsilon_t)
  \, d\varepsilon_t \, dx_{t-1}.
  \label{eq:bf-structural-ukf-prop-pushforward}
\end{equation}
Equation~\eqref{eq:bf-structural-ukf-prop-pushforward} is exactly the
expectation of $\varphi$ under the pushforward of the joint law of
$(x_{t-1},\varepsilon_t)$ through $F$, which proves part~(i).

Moreover, under the structural transition,
\begin{equation}
  x_t = F(x_{t-1},\varepsilon_t)
  \qquad \text{almost surely.}
  \label{eq:bf-structural-ukf-prop-deterministic-completion}
\end{equation}
So once $(x_{t-1},\varepsilon_t)$ is fixed, both $m_t$ and $k_t$ are fixed.
There is no independent new perturbation attached to $k_t$ under the intended
model.  Therefore any sigma-point rule intended to approximate
$p(x_t\mid y_{1:t-1})$ should be applied to the pre-transition uncertainty
variables that generate that law, namely $(x_{t-1},\varepsilon_t)$ or an
exactly equivalent parameterization.  This proves part~(ii).

Finally, if a full-state additive-noise UKF omits $\varepsilon_t$, then it
misses genuine current-shock uncertainty.  If it instead adds a direct
perturbation to $k_t$, then it enlarges the stochastic space and defines a
different predictive law unless that perturbation is degenerate in exactly the
way required to reproduce the same pushforward distribution in
\eqref{eq:bf-structural-ukf-prop-pushforward}.  This proves part~(iii).
\end{proof}

The proposition does not say that the filter stops targeting $x_t$.  The
filtered state remains $x_t$.  The point is that the sigma-point variable is the
pre-transition uncertainty whose pushforward generates the predictive law of
$x_t$.  This is analogous to process-noise augmentation in a standard additive-
noise UKF, but in the mixed structural setting the distinction is sharper:
$\varepsilon_t$ is a genuine new source of uncertainty, while $k_t$ is a
structural completion of that uncertainty rather than an independently shocked
coordinate.

\subsection{Comparison Table}
\label{sec:bf-structural-ukf-comparison-table}

\begin{center}
\begin{tabular}{p{0.18\linewidth}p{0.34\linewidth}p{0.34\linewidth}}
\toprule
\textbf{Feature} & \textbf{Structural UKF} & \textbf{Naive full-state UKF}\\
\midrule
Integration space & Pre-transition variables, e.g. $(x_{t-1},\varepsilon_t)$ or the declared stochastic block & Full post-transition state with additive covariance on all coordinates\\
Independent new randomness & Only in the declared stochastic block & Potentially assigned to every state coordinate\\
Deterministic-block treatment & Completed pointwise from the structural map & Perturbed directly unless extra constraints are imposed\\
Support preservation & Preserves the support implied by the structural transition & Can place mass outside that support\\
Target law & Same structural target the model declares & Different target unless justified as an approximation\\
Metadata label & No approximation label when faithful to the contract & Approximation label required if the stochastic space is enlarged\\
When acceptable & Default design for mixed structural transitions & Only as a declared and tested approximation\\
\bottomrule
\end{tabular}
\end{center}

Let
\begin{align}
  m_t &= \rho m_{t-1}+\sigma\varepsilon_t,
    & \varepsilon_t &\sim \calN(0,1), \label{eq:bf-ukf-example-m}\\
  k_t &= \phi k_{t-1}+\gamma m_t^2, \label{eq:bf-ukf-example-k}\\
  y_t &= m_t+k_t+e_t,
    & e_t &\sim \calN(0,R). \label{eq:bf-ukf-example-y}
\end{align}
The state is $x_t=(m_t,k_t)$, but only the innovation
$\varepsilon_t$ is new stochastic input at time $t$.  Conditional on
$(m_{t-1},k_{t-1},\varepsilon_t)$, the coordinate $k_t$ is deterministic.

Suppose the previous filtering distribution is Gaussian:
\[
  x_{t-1}\mid y_{1:t-1}
  \approx \calN(\mu_{t-1},P_{t-1}),
  \qquad
  \mu_{t-1}=
  \begin{bmatrix}\bar m\\ \bar k\end{bmatrix}.
\]
The structural UKF should augment the previous state with the innovation.  This
step needs to be motivated carefully, because it is exactly where the structural
UKF departs from the naive full-state construction.  The point is not to pad the
state arbitrarily.  The point is to place sigma points on the variables that
actually generate the one-step uncertainty in the structural transition.  In
this example, the predictive law is generated by the joint uncertainty in the
lagged state $(m_{t-1},k_{t-1})$ and the current shock $\varepsilon_t$.  Once
those three quantities are fixed, both $m_t$ and $k_t$ are fixed by the
structural map.

That is why the augmented variable is
\begin{equation}
  a_t=(x_{t-1},\varepsilon_t),
  \label{eq:bf-ukf-augmented-variable}
\end{equation}
rather than an artificially padded post-transition full state.  Standard
additive-noise UKF presentations often augment with process noise for the same
reason: sigma points should live on the pre-transition uncertainty variables.
For a mixed structural model, however, that principle has a sharper consequence.
The shock $\varepsilon_t$ is a genuine new source of uncertainty, while $k_t$
is not.  If the augmentation omitted $\varepsilon_t$, the propagated sigma
points would miss current shock uncertainty.  If the augmentation instead added
a direct perturbation for $k_t$, the sigma points would describe a different
transition law from the one declared by the model.

So the structural UKF should augment the previous state with the innovation.
This places sigma points in the same pre-transition uncertainty variables that
define the predictive law, rather than in an artificially padded post-transition
full state:
\begin{equation}
  a_t =
  \begin{bmatrix}
    m_{t-1}\\ k_{t-1}\\ \varepsilon_t
  \end{bmatrix},
  \qquad
  a_t\mid y_{1:t-1}
  \approx
  \calN\left(
    \begin{bmatrix}\bar m\\ \bar k\\ 0\end{bmatrix},
    \begin{bmatrix}
      P_{t-1} & 0\\
      0 & 1
    \end{bmatrix}
  \right).
  \label{eq:bf-ukf-augmented-law}
\end{equation}
Let $\{a_t^{(j)},w_j^{(m)},w_j^{(c)}\}_{j=0}^{2n_a}$ be the usual unscented
points and weights for this $n_a=3$ dimensional augmented Gaussian
\citep{julier1997new}.  For every point
$a_t^{(j)}=(m_{t-1}^{(j)},k_{t-1}^{(j)},\varepsilon_t^{(j)})$, the structural
transition is evaluated as
\begin{align}
  m_t^{(j)}
    &= \rho m_{t-1}^{(j)}+\sigma\varepsilon_t^{(j)},\\
  k_t^{(j)}
    &= \phi k_{t-1}^{(j)}+\gamma\left(m_t^{(j)}\right)^2,\\
  x_t^{(j)}
    &= \begin{bmatrix}m_t^{(j)}\\ k_t^{(j)}\end{bmatrix}.
\end{align}
This is the crucial step: the filter never creates an independent sigma-point
perturbation for $k_t$.  It creates sigma points for the previous state and
the current innovation, then computes $k_t$ by the structural map
\eqref{eq:bf-ukf-example-k}.

The structural and naive full-state UKF constructions now differ at the level
that matters for the update.  The structural UKF forms sigma points for
$(x_{t-1},\varepsilon_t)$, propagates them through the structural transition,
and therefore computes moments of the target law declared by the model.  A
naive full-state UKF instead behaves as if there were an additive-noise model
of the form
\[
  x_t = \tilde f(x_{t-1}) + \tilde u_t,
  \qquad
  \tilde u_t \sim \calN(0,\tilde Q),
\]
with a full-state covariance $\tilde Q$ that assigns direct new variance to all
or most coordinates of $x_t$.  The three failure claims can now be derived in
the notation of this toy model.

\paragraph{Prediction-law failure.}
Under the structural model,
\[
  m_t = \rho m_{t-1} + \sigma \varepsilon_t,
  \qquad
  k_t = \phi k_{t-1} + \gamma m_t^2.
\]
So conditional on $(m_{t-1},k_{t-1})$, the one-step predictive law is induced by
a single scalar shock $\varepsilon_t$.  Eliminating the shock-driven update for
$m_t$ gives the structural identity
\begin{equation}
  k_t - \phi k_{t-1} - \gamma m_t^2 = 0.
  \label{eq:bf-ukf-structural-identity}
\end{equation}
Every propagated sigma point in the structural UKF satisfies this identity, so
its sigma-point cloud lies on the support of the structural predictive law.  A
naive full-state UKF instead propagates points
\[
  \tilde x_t^{(j)}=(\tilde m_t^{(j)},\tilde k_t^{(j)})
\]
from an additive full-state covariance model.  In general those points satisfy
\[
  \tilde k_t^{(j)} - \phi k_{t-1}^{(j)} - \gamma \bigl(\tilde m_t^{(j)}\bigr)^2
  \neq 0,
\]
so the naive sigma-point cloud approximates moments of a different one-step law.
That is the prediction-law failure.

\paragraph{Cross-covariance failure.}
For the observation equation $z_t=m_t+k_t$, the structural UKF computes
\[
  P_{xz,t}^{\mathrm{struct}}
  =
  \sum_j w_j^{(c)}
  \bigl(x_t^{(j)}-\hat x_{t|t-1}\bigr)
  \bigl(z_t^{(j)}-\hat z_t\bigr),
\]
with
\[
  z_t^{(j)} = m_t^{(j)} + k_t^{(j)},
  \qquad
  k_t^{(j)} = \phi k_{t-1}^{(j)} + \gamma \bigl(m_t^{(j)}\bigr)^2.
\]
Hence all variation in the second state coordinate enters through lagged-state
variation and the common shock channel already represented in $m_t^{(j)}$.
Under the naive full-state UKF,
\[
  P_{xz,t}^{\mathrm{naive}}
  =
  \sum_j w_j^{(c)}
  \bigl(\tilde x_t^{(j)}-\tilde x_{t|t-1}\bigr)
  \bigl(\tilde z_t^{(j)}-\tilde z_t\bigr),
\]
with
\[
  \tilde z_t^{(j)} = \tilde m_t^{(j)} + \tilde k_t^{(j)}.
\]
Because $\tilde k_t^{(j)}$ now contains a direct artificial perturbation channel,
the second component of $P_{xz,t}^{\mathrm{naive}}$ contains covariance terms
between the observation and that artificial variation.  Those terms do not exist
in the structural model, where $k_t$ is completed deterministically from the
same shock path that drives $m_t$.  That is the cross-covariance failure.

\paragraph{Update-law failure.}
The familiar UKF update equations,
\[
  K_t = P_{xz,t} S_t^{-1},
  \qquad
  \hat x_{t|t} = \hat x_{t|t-1} + K_t v_t,
  \qquad
  P_{t|t} = P_{t|t-1} - K_t S_t K_t^\top,
\]
do not fail algebraically.  They fail semantically when
$(\hat x_{t|t-1},S_t,P_{xz,t})$ are computed from the wrong predictive law.
Once the naive full-state approximation changes either $P_{xz,t}$ or $S_t$, it
changes the gain itself:
\[
  K_t^{\mathrm{naive}}
  =
  P_{xz,t}^{\mathrm{naive}} \left(S_t^{\mathrm{naive}}\right)^{-1}
  \neq
  P_{xz,t}^{\mathrm{struct}} \left(S_t^{\mathrm{struct}}\right)^{-1}
  = K_t^{\mathrm{struct}}
\]
in general.  The existing numerical comparison later in this example shows this
explicitly: adding artificial variance to $k_t$ changes the innovation variance
from $S_t=0.61217$ to $S_t=0.65217$, which then changes the approximate
log-likelihood contribution.  The same change propagates into the Kalman gain
and posterior update.  So the update law is not wrong as linear algebra; it is
correctly updating the wrong approximate model.

The predicted state moments are
\begin{align}
  \hat x_{t|t-1}
    &= \sum_j w_j^{(m)} x_t^{(j)},\\
  P_{xx,t}
    &= \sum_j w_j^{(c)}
       \left(x_t^{(j)}-\hat x_{t|t-1}\right)
       \left(x_t^{(j)}-\hat x_{t|t-1}\right)^\top .
\end{align}
For the observation equation \eqref{eq:bf-ukf-example-y}, define
\[
  z_t^{(j)}=m_t^{(j)}+k_t^{(j)}.
\]
Then the predicted observation, innovation variance, and cross covariance are
\begin{align}
  \hat z_t
    &= \sum_j w_j^{(m)} z_t^{(j)},\\
  S_t
    &= \sum_j w_j^{(c)}(z_t^{(j)}-\hat z_t)^2 + R,\\
  P_{xz,t}
    &= \sum_j w_j^{(c)}
       \left(x_t^{(j)}-\hat x_{t|t-1}\right)(z_t^{(j)}-\hat z_t).
\end{align}
The Gaussian measurement update is
\begin{align}
  v_t &= y_t-\hat z_t,\\
  K_t &= P_{xz,t}S_t^{-1},\\
  \hat x_{t|t} &= \hat x_{t|t-1}+K_tv_t,\\
  P_{t|t} &= P_{xx,t}-K_tS_tK_t^\top,
\end{align}
and the one-step approximate log likelihood contribution is
\[
  \ell_t
  =
  -\frac{1}{2}\left[
    \log(2\pi S_t)+\frac{v_t^2}{S_t}
  \right].
\]

\subsection{Numerical Illustration}
\label{sec:bf-structural-ukf-numeric}

Use the parameters
\[
  \rho=0.8,\quad \sigma=0.5,\quad \phi=0.7,\quad
  \gamma=0.4,\quad R=0.25,
\]
and previous filtering moments
\[
  \mu_{t-1}=(0,0)^\top,
  \qquad
  P_{t-1}=\operatorname{diag}(0.04,0.09).
\]
With the scaled unscented choice $\lambda=0$, the mean weight of the central
point is zero and the six off-center points each have mean weight $1/6$.
The augmented covariance is
\[
  \operatorname{diag}(0.04,0.09,1),
\]
so $\sqrt{(n_a+\lambda)P_a}$ has diagonal entries
$0.34641$, $0.51962$, and $1.73205$.  The seven sigma points and their
deterministic structural completion are:
\[
\begin{array}{c|rrr|rrr}
  j
  & m_{t-1}^{(j)} & k_{t-1}^{(j)} & \varepsilon_t^{(j)}
  & m_t^{(j)} & k_t^{(j)} & z_t^{(j)}\\
  \hline
  0 & 0       & 0        & 0        & 0        & 0        & 0\\
  1 & 0.34641& 0        & 0        & 0.27713 & 0.03072 & 0.30785\\
  2 & 0      & 0.51962  & 0        & 0       & 0.36373 & 0.36373\\
  3 & 0      & 0        & 1.73205  & 0.86603 & 0.30000 & 1.16603\\
  4 & -0.34641& 0       & 0        & -0.27713& 0.03072 & -0.24641\\
  5 & 0      & -0.51962 & 0        & 0       & -0.36373& -0.36373\\
  6 & 0      & 0        & -1.73205 & -0.86603& 0.30000 & -0.56603
\end{array}
\]
For example, the third positive innovation point gives
$m_t=0.5(1.73205)=0.86603$ and
$k_t=0.4(0.86603)^2=0.30000$.  No row contains an independently sampled
``new'' shock for $k_t$.

Using covariance weight $w_0^{(c)}=2$ for the central point, the structural
propagation gives
\[
  \hat x_{t|t-1}
  =
  \begin{bmatrix}
    0\\ 0.11024
  \end{bmatrix},
  \qquad
  P_{xx,t}
  =
  \begin{bmatrix}
    0.27560 & 0\\
    0 & 0.08657
  \end{bmatrix}.
\]
The predicted observation moments are
\[
  \hat z_t=0.11024,\qquad S_t=0.61217,\qquad
  P_{xz,t}=
  \begin{bmatrix}
    0.27560\\ 0.08657
  \end{bmatrix}.
\]
For an observation $y_t=0.30$,
\[
  v_t=0.18976,\qquad
  K_t=
  \begin{bmatrix}
    0.45020\\ 0.14141
  \end{bmatrix},
\]
and therefore
\[
  \hat x_{t|t}
  =
  \begin{bmatrix}
    0.08543\\ 0.13707
  \end{bmatrix},
  \qquad
  P_{t|t}
  =
  \begin{bmatrix}
    0.15152 & -0.03897\\
    -0.03897 & 0.07433
  \end{bmatrix}.
\]
The one-step log likelihood contribution is approximately
\[
  \ell_t=-0.70297.
\]

\paragraph{Side-by-side prediction/update contrast.}
For this toy model, the structural UKF treats
\[
  a_t=(x_{t-1},\varepsilon_t)
\]
as the sigma-point object and computes propagated points by
\[
  m_t^{(j)} = \rho m_{t-1}^{(j)} + \sigma \varepsilon_t^{(j)},
  \qquad
  k_t^{(j)} = \phi k_{t-1}^{(j)} + \gamma \bigl(m_t^{(j)}\bigr)^2.
\]
It therefore forms
\[
  P_{xz,t}^{\mathrm{struct}}
  =
  \sum_j w_j^{(c)}
  \bigl(x_t^{(j)}-\hat x_{t|t-1}\bigr)
  \bigl(z_t^{(j)}-\hat z_t\bigr),
\]
where every $x_t^{(j)}$ lies on the support generated by the structural map.
A naive full-state UKF instead produces an artificial cloud
$\tilde x_t^{(j)}=(\tilde m_t^{(j)},\tilde k_t^{(j)})$ with
\[
  \tilde k_t^{(j)} \neq \phi k_{t-1}^{(j)} + \gamma \bigl(\tilde m_t^{(j)}\bigr)^2
  \text{ in general,}
\]
and therefore forms
\[
  P_{xz,t}^{\mathrm{naive}}
  =
  \sum_j w_j^{(c)}
  \bigl(\tilde x_t^{(j)}-\tilde x_{t|t-1}\bigr)
  \bigl(\tilde z_t^{(j)}-\tilde z_t\bigr),
\]
which mixes genuine shock-driven variation with artificial perturbations of the
deterministic-completion coordinate.

\paragraph{Structural identity test.}
A simple way to see the difference is to check whether propagated points satisfy
\[
  k_t - \phi k_{t-1} - \gamma m_t^2 = 0.
\]
The structural UKF satisfies this identity point by point for every propagated
sigma point.  A naive full-state UKF with direct perturbations to $k_t$ does
not.  That is the exact place where the structural target and the naive
full-state approximation part company.

Now compare this with an off-manifold full-state approximation that adds an
independent artificial innovation variance $0.04$ directly to $k_t$ after the
structural map.  Nothing in
\eqref{eq:bf-ukf-example-m}--\eqref{eq:bf-ukf-example-y} justifies that extra
shock.  It changes the innovation variance from $S_t=0.61217$ to
$S_t=0.65217$ and changes the same observation's log likelihood contribution
from $-0.70297$ to $-0.73282$.  The numerical difference is small in this toy
calibration, but the conceptual difference is large: the latter filter is
approximating a different state-space model.

To connect these numbers back to the three failures above, write the naive
full-state approximation as using
\[
  \tilde P_{xx,t}
  =
  \begin{bmatrix}
    0.27560 & 0\\
    0 & 0.12657
  \end{bmatrix},
\]
which differs from the structural covariance only by the artificial extra
variance in the $k_t$ coordinate.  Under the same observation equation
$z_t=m_t+k_t$, this gives
\[
  \tilde S_t = 0.65217,
  \qquad
  \tilde P_{xz,t} =
  \begin{bmatrix}
    0.27560\\ 0.12657
  \end{bmatrix},
\]
and therefore
\[
  \tilde K_t
  =
  \begin{bmatrix}
    0.42259\\ 0.19408
  \end{bmatrix},
  \qquad
  \tilde x_{t|t}
  =
  \begin{bmatrix}
    0.08019\\ 0.14707
  \end{bmatrix},
\]
for the same observation $y_t=0.30$.

These values make the three failures concrete:
\begin{enumerate}[label=(\roman*)]
  \item \textbf{Prediction-law failure:}
    the structural update uses
    $P_{xx,t}=\operatorname{diag}(0.27560,0.08657)$,
    while the naive full-state approximation uses
    $\tilde P_{xx,t}=\operatorname{diag}(0.27560,0.12657)$ because it has
    inserted artificial uncertainty directly into $k_t$.
  \item \textbf{Cross-covariance failure:}
    the structural update uses
    \[
      P_{xz,t}=
      \begin{bmatrix}
        0.27560\\ 0.08657
      \end{bmatrix},
    \]
    whereas the naive full-state approximation uses
    \[
      \tilde P_{xz,t}=
      \begin{bmatrix}
        0.27560\\ 0.12657
      \end{bmatrix}.
    \]
    The second component is larger only because the naive construction lets the
    observation load directly on artificial variation in $k_t$.
  \item \textbf{Update-law failure:}
    the structural gain and posterior mean are
    \[
      K_t=
      \begin{bmatrix}
        0.45020\\ 0.14141
      \end{bmatrix},
      \qquad
      \hat x_{t|t}=
      \begin{bmatrix}
        0.08543\\ 0.13707
      \end{bmatrix},
    \]
    whereas the naive full-state approximation gives
    \[
      \tilde K_t=
      \begin{bmatrix}
        0.42259\\ 0.19407
      \end{bmatrix},
      \qquad
      \tilde x_{t|t}=
      \begin{bmatrix}
        0.08019\\ 0.14707
      \end{bmatrix}.
    \]
    The update formula has not changed, but the gain and posterior correction do
    because they are now computed from the wrong predictive moments.
\end{enumerate}
So the numerical illustration does not merely show a slightly different
likelihood value.  It shows that once the naive approximation changes the
predictive covariance and cross covariance, the entire measurement update shifts
with it.

The lesson is not that UKF is wrong.  The lesson is that the UKF must be
applied to the correct structural transition.  Sigma points belong in the
declared stochastic integration space.  Endogenous deterministic coordinates
are completed by the model point by point.  In BayesFilter terms, this is a
design rule for mixed structural models and an approximation-label boundary for
any backend that instead enlarges the stochastic space.

\section{Nonlinear Measurement Equations with a Degenerate Linear Transition}
\label{sec:bf-structural-nonlinear-measurement}

The worked UKF example above uses a nonlinear state transition.  Many applied
state-space problems instead have a degenerate linear transition together with a
nonlinear measurement equation.  That case is closer to many DSGE and affine
term-structure applications than a fully nonlinear latent transition.  The key
point is that the latent transition may still have a structural split even when
it is linear.  A nonlinear measurement equation does not remove that split.

Consider the linear structural transition
\begin{align}
  m_t &= \rho m_{t-1} + \sigma\varepsilon_t,
    & \varepsilon_t &\sim \calN(0,1),
    \label{eq:bf-exp-affine-transition-m}\\
  k_t &= \phi k_{t-1} + \psi m_{t-1} + \chi m_t,
    \label{eq:bf-exp-affine-transition-k}\\
  x_t &= (m_t,k_t),
    \label{eq:bf-exp-affine-transition-state}
\end{align}
and the nonlinear exponential-affine measurement equation
\begin{align}
  y_t &= \exp(A^\top x_t + \mu) + e_t,
    & e_t &\sim \calN(0,R).
    \label{eq:bf-exp-affine-observation}
\end{align}
The latent transition is linear, but it is still structurally degenerate:
conditional on $(x_{t-1},\varepsilon_t)$, the endogenous coordinate $k_t$ is
completed deterministically from the same shock path that drives $m_t$.  So the
structural issue from the previous section does not disappear.  What changes is
where the nonlinearity appears.  The latent transition is linear, while the
measurement map is nonlinear.

This means there are now two layers to keep separate:
\begin{enumerate}[label=(\roman*)]
  \item the latent predictive law is still generated by the pre-transition
    uncertainty variables $(x_{t-1},\varepsilon_t)$, not by direct full-state
    perturbations of $x_t$;
  \item even after the latent cloud is propagated correctly, the UKF must still
    approximate nonlinear observation moments and cross covariances for the map
    $x_t \mapsto \exp(A^\top x_t + \mu)$.
\end{enumerate}
The first issue is structural-modeling correctness.  The second is observation-
side quadrature quality.

\subsection{Formal structural statement for the degenerate linear-transition case}
\label{sec:bf-exp-affine-formal}

\paragraph{Accepted assumptions.}
For this section, treat the following as explicit modeling assumptions:
\begin{enumerate}[label=(A\arabic*)]
  \item the innovation is conditionally independent of the lagged filtering law,
    so that
    \begin{equation}
      p(x_{t-1},\varepsilon_t\mid y_{1:t-1})
      =
      p(x_{t-1}\mid y_{1:t-1})p(\varepsilon_t);
      \label{ass:bf-exp-affine-factorization}
    \end{equation}
  \item the latent map, observation map, and composed map defined below are
    measurable and have the finite moments needed for the stated expectations;
  \item whenever we refer to the observation-side-only error regime, we assume
    that the latent predictive law from
    \eqref{eq:bf-exp-affine-pushforward} is preserved exactly.
\end{enumerate}

\begin{proposition}[Latent predictive pushforward]
\label{lem:bf-exp-affine-latent-pushforward}
Define the degenerate linear structural map
\begin{equation}
  F_{\mathrm{lin}}(x_{t-1},\varepsilon_t)
  =
  \Bigl(
    \rho m_{t-1}+\sigma\varepsilon_t,
    \phi k_{t-1}+\psi m_{t-1}+\chi(\rho m_{t-1}+\sigma\varepsilon_t)
  \Bigr).
  \label{eq:bf-exp-affine-latent-map}
\end{equation}
Then for any bounded test function $\varphi$ on the latent-state space,
\begin{equation}
  \E\!\left[\varphi(x_t)\mid y_{1:t-1}\right]
  =
  \iint
  \varphi\!\left(F_{\mathrm{lin}}(x_{t-1},\varepsilon_t)\right)
  p(x_{t-1}\mid y_{1:t-1})p(\varepsilon_t)
  \, d\varepsilon_t \, dx_{t-1}.
  \label{eq:bf-exp-affine-pushforward}
\end{equation}
Hence the latent predictive law is the pushforward of the joint law of
$(x_{t-1},\varepsilon_t)$ through $F_{\mathrm{lin}}$.
\end{proposition}

\begin{proof}
Under the state transition
\eqref{eq:bf-exp-affine-transition-m}--\eqref{eq:bf-exp-affine-transition-state}
and the factorization assumption
\eqref{ass:bf-exp-affine-factorization}, the conditional expectation of any
bounded test function $\varphi(x_t)$ is obtained by substituting
$x_t=F_{\mathrm{lin}}(x_{t-1},\varepsilon_t)$ and integrating over the joint law
of $(x_{t-1},\varepsilon_t)$.  This gives
\eqref{eq:bf-exp-affine-pushforward}, which is exactly the pushforward identity.
\end{proof}

\begin{proposition}[Deterministic-completion identity]
\label{lem:bf-exp-affine-deterministic-completion}
Under \eqref{eq:bf-exp-affine-transition-m}--\eqref{eq:bf-exp-affine-transition-k},
all admissible latent states satisfy
\begin{equation}
  k_t - \phi k_{t-1} - \psi m_{t-1} - \chi m_t = 0.
  \label{eq:bf-exp-affine-structural-identity}
\end{equation}
Therefore, once $(x_{t-1},\varepsilon_t)$ is fixed, both $m_t$ and $k_t$ are
fixed; there is no independent new perturbation attached to $k_t$ in the latent
transition.
\end{proposition}

\begin{proof}
Substitute \eqref{eq:bf-exp-affine-transition-m} into
\eqref{eq:bf-exp-affine-transition-k} and rearrange terms.  The resulting
identity is exactly \eqref{eq:bf-exp-affine-structural-identity}.
\end{proof}

\begin{proposition}[Observation predictive pushforward]
\label{lem:bf-exp-affine-observation-pushforward}
Define the nonlinear noise-free observation map
\begin{equation}
  h(x_t)=\exp(A^\top x_t+\mu).
  \label{eq:bf-exp-affine-observation-map}
\end{equation}
and the composed map
\begin{equation}
  G(x_{t-1},\varepsilon_t)
  =
  h\!\left(F_{\mathrm{lin}}(x_{t-1},\varepsilon_t)\right).
  \label{eq:bf-exp-affine-composed-map}
\end{equation}
Then for any bounded test function $\psi_0$ on the noise-free observation
space,
\begin{equation}
  \E\!\left[\psi_0\!\bigl(h(x_t)\bigr)\mid y_{1:t-1}\right]
  =
  \iint
  \psi_0\!\left(G(x_{t-1},\varepsilon_t)\right)
  p(x_{t-1}\mid y_{1:t-1})p(\varepsilon_t)
  \, d\varepsilon_t \, dx_{t-1}.
  \label{eq:bf-exp-affine-observation-pushforward}
\end{equation}
Hence the noise-free observation predictive law is the pushforward of the same
pre-transition uncertainty through the composed map.
\end{proposition}

\begin{proof}
Compose the latent pushforward identity
\eqref{eq:bf-exp-affine-pushforward} with the measurable observation map
\eqref{eq:bf-exp-affine-observation-map}.  This yields
\eqref{eq:bf-exp-affine-observation-pushforward}.
\end{proof}

\begin{proposition}[Naive full-state perturbation changes the latent law]
\label{lem:bf-exp-affine-naive-full-state-support-violation}
If a naive full-state UKF introduces a direct perturbation to the endogenous
coordinate $k_t$, then its propagated latent sigma-point cloud does not, in
general, satisfy \eqref{eq:bf-exp-affine-structural-identity} pointwise.
Therefore it does not target the latent predictive law characterized by
\eqref{eq:bf-exp-affine-pushforward}.
\end{proposition}

\begin{proof}
By Lemma~\ref{lem:bf-exp-affine-deterministic-completion}, every admissible
latent state under the structural model satisfies
\eqref{eq:bf-exp-affine-structural-identity}.  A naive full-state UKF that adds
an independent perturbation to $k_t$ creates propagated points whose $k_t$
values are not determined solely by $(x_{t-1},\varepsilon_t)$, so the identity
fails in general.  Hence the resulting sigma-point cloud cannot represent the
same latent predictive law.
\end{proof}

\begin{proposition}[Failure propagation and observation-side-only error]
\label{prop:bf-exp-affine-law-mismatch-propagation}
Under the preceding lemmas:
\begin{enumerate}[label=(\roman*)]
  \item if a naive full-state UKF directly perturbs $k_t$, then
    prediction-law failure, cross-covariance failure, and update-law failure all
    still apply;
  \item if the latent predictive law from
    \eqref{eq:bf-exp-affine-pushforward} is preserved exactly and only the
    transformed observation moments from
    \eqref{eq:bf-exp-affine-observation-pushforward} are approximated by
    quadrature, then the error is not structural-model failure; it is only
    observation-side quadrature error.
\end{enumerate}
\end{proposition}

\begin{proof}
Part~(i) follows because Lemma~\ref{lem:bf-exp-affine-naive-full-state-support-violation}
shows that the naive construction changes the latent predictive law, and
Lemma~\ref{lem:bf-exp-affine-observation-pushforward} shows that the nonlinear
observation law is the pushforward of that same latent law.  So once the latent
law changes, the observation moments, state-observation cross covariance, and
therefore the UKF update objects also change.  Part~(ii) follows from Assumption
(A3): if the latent law is preserved and only the nonlinear observation moments
are approximated, then the remaining error arises only in the observation-side
quadrature.
\end{proof}

The formal statements above separate three layers that are easy to blur in
practice: accepted modeling assumptions, exact pushforward identities, and the
interpretation of failure modes.  The filtered state is still $x_t$, but the
latent predictive law is generated by the pre-transition uncertainty variables
$(x_{t-1},\varepsilon_t)$.  The nonlinear measurement equation adds another
approximation layer, but it does not remove the need to respect the degenerate
latent transition.

\subsection{Worked Degenerate Linear-Transition Example}
\subsection{Worked Degenerate Linear-Transition Example}
\label{sec:bf-exp-affine-example}

Use
\[
  \rho = 0.8,
  \qquad
  \sigma = 0.5,
  \qquad
  \phi = 0.7,
  \qquad
  \psi = 0.3,
  \qquad
  \chi = 0.4,
\]
with previous filtering moments
\[
  \mu_{t-1}=(0,0)^\top,
  \qquad
  P_{t-1}=\operatorname{diag}(0.04,0.09),
\]
and measurement parameters
\[
  A=
  \begin{bmatrix}
    1.1\\ 0.6
  \end{bmatrix},
  \qquad
  \mu = 0.10,
  \qquad
  R = 0.04.
\]
Because the latent transition remains structurally degenerate, the correct
sigma-point object is still
\[
  a_t=(x_{t-1},\varepsilon_t),
\]
not a naively perturbed full post-transition state.  Using the same scaled
unscented construction as in the previous example, the propagated latent points
are obtained from
\begin{align}
  m_t^{(j)} &= \rho m_{t-1}^{(j)} + \sigma\varepsilon_t^{(j)},\\
  k_t^{(j)} &= \phi k_{t-1}^{(j)} + \psi m_{t-1}^{(j)} + \chi m_t^{(j)}.
\end{align}
For each propagated point, define the nonlinear noise-free observation
\[
  z_t^{(j)} = \exp\!\left(1.1 m_t^{(j)} + 0.6 k_t^{(j)} + 0.10\right).
\]
The same seven sigma points used earlier now produce the following latent and
measurement values:
\[
\begin{array}{c|rrr|rrr}
  j
  & m_{t-1}^{(j)} & k_{t-1}^{(j)} & \varepsilon_t^{(j)}
  & m_t^{(j)} & k_t^{(j)} & z_t^{(j)}\\
  \hline
  0 & 0 & 0 & 0 & 0 & 0 & 1.10517\\
  1 & 0.34641 & 0 & 0 & 0.27713 & 0.21455 & 1.61857\\
  2 & 0 & 0.51962 & 0 & 0 & 0.36373 & 1.37758\\
  3 & 0 & 0 & 1.73205 & 0.86603 & 0.34641 & 3.39053\\
  4 & -0.34641 & 0 & 0 & -0.27713 & -0.21455 & 0.75442\\
  5 & 0 & -0.51962 & 0 & 0 & -0.36373 & 0.88816\\
  6 & 0 & 0 & -1.73205 & -0.86603 & -0.34641 & 0.36079
\end{array}
\]
The predicted latent moments are
\[
  \hat x_{t|t-1}
  =
  \begin{bmatrix}
    0\\ 0
  \end{bmatrix},
  \qquad
  P_{xx,t}
  =
  \begin{bmatrix}
    0.27560 & 0.16536\\
    0.16536 & 0.12170
  \end{bmatrix}.
\]
The nonlinear predicted observation moments are
\[
  \hat z_t = 1.39834,
  \qquad
  S_t = 1.08356,
  \qquad
  P_{xz,t}
  =
  \begin{bmatrix}
    0.39302\\ 0.24514
  \end{bmatrix}.
\]
So the structural UKF gain is
\[
  K_t
  =
  \begin{bmatrix}
    0.36270\\ 0.22624
  \end{bmatrix}.
\]
If the realized observation is $y_t=1.50$, then
\[
  v_t = 0.10166,
  \qquad
  \hat x_{t|t}
  =
  \begin{bmatrix}
    0.03687\\ 0.02300
  \end{bmatrix}.
\]

\subsection{What Is Similar to the Structural-Deterministic Case?}
\label{sec:bf-exp-affine-similarities}

The similarity is that the structural latent transition is still generated by
$(x_{t-1},\varepsilon_t)$, not by independent perturbations of every coordinate
of $x_t$.  If a naive full-state UKF directly perturbs $k_t$, it again changes
the latent predictive law before the observation map is even evaluated.  So the
structural warning remains active even though the latent transition is linear.

The nonlinear observation adds a second approximation layer.  Once the latent
cloud is propagated correctly, the UKF still has to approximate
\[
  \hat y_t,
  \qquad
  S_t,
  \qquad
  P_{xz,t}
\]
for the map $x_t \mapsto \exp(A^\top x_t + \mu)$.  Poor quadrature here still
changes the gain and posterior update, just as in other nonlinear Gaussian
filters.

\subsection{What Is Different from the Earlier Nonlinear-State Example?}
\label{sec:bf-exp-affine-differences}

In the earlier toy example, the latent state equation itself was nonlinear, so
the UKF had to approximate both the latent pushforward through the nonlinear
state map and the resulting observation moments.  Here the latent transition is
linear, but structurally degenerate.  That removes one source of approximation
error but not the structural state-partition issue.  The UKF still needs the
correct pre-transition uncertainty variable because $k_t$ remains a
deterministic-completion coordinate.

So this example isolates a different practical regime:
\begin{enumerate}[label=(\roman*)]
  \item the latent transition is linear but not full-state stochastic;
  \item the nonlinear burden sits in the measurement equation;
  \item the UKF can still fail in two distinct ways:
    \begin{itemize}
      \item structurally, if it perturbs $k_t$ directly and therefore changes the
        latent predictive law;
      \item numerically, if it uses the correct latent construction but poorly
        approximates the nonlinear observation moments.
    \end{itemize}
\end{enumerate}
This is exactly why the structural deterministic-dynamics chapter should not be
read as applying only to nonlinear state transitions.  The same latent
state-partition discipline remains necessary when the state equation is linear
but degenerate and the observation map is nonlinear.

\section{Pruned DSGE Filtering: A Minimal Correct Pattern}
\label{sec:bf-pruned-dsge-correct-pattern}

For pruned second-order DSGE approximations, the source DSGE project already
uses the standard first-order/second-order decomposition:
\begin{align}
  x_t^f &= h_x x_{t-1}^f + \eta\varepsilon_t,\\
  x_t^s &= h_x x_{t-1}^s
    + \frac{1}{2}H_{xx}(x_{t-1}^f\otimes x_{t-1}^f)
    + \frac{1}{2}h_{ss},\\
  y_t &= g_x(x_t^f+x_t^s)
    + \frac{1}{2}G_{xx}(x_t^f\otimes x_t^f)
    + \frac{1}{2}g_{ss} + e_t .
\end{align}
That decomposition is necessary, but it is not sufficient for large DSGE
filtering.  It separates perturbation order; it does not automatically
separate exogenous shock states from endogenous deterministic states inside
$x_t^f$.  A robust BayesFilter DSGE adapter should carry both decompositions:
\[
  \text{perturbation order: } (x^f,x^s),
  \qquad
  \text{structural timing: } (m,k).
\]

The implementation should therefore expose a structural transition interface
with at least:
\begin{itemize}
  \item exogenous-state indices and shock dimensions;
  \item endogenous/predetermined deterministic-state indices;
  \item the exogenous transition $T_m$;
  \item the deterministic endogenous transition $T_k$;
  \item the observation map and measurement-error law;
  \item a declaration of whether the backend integrates over shocks,
    exogenous states, or a full latent state approximation.
\end{itemize}
For CKF/UKF/SVD sigma-point filters, sigma points should be generated in the
declared stochastic integration space and then lifted through the structural
transition.  For particle filters, proposals should sample the stochastic
blocks and deterministically complete the endogenous block unless a different
proposal is explicitly specified.

\section{Current Source-Project Failure Mode}
\label{sec:bf-current-source-project-failure-mode}

The recent source-project audit found a concrete instance of this design
failure.  The default DSGE SVD sigma-point adapter uses a collapsed transition
of the form
\[
  Q_w = \eta\eta^\top + \epsilon I,\qquad
  x_{t|t-1}^f = h_x x_{t-1|t-1}^f,
\]
and propagates the covariance as
\[
  P_{t|t-1} = h_x P_{t-1|t-1} h_x^\top + Q_w.
\]
It tracks the pruned $x^s$ correction separately, but it does not expose a
partition between exogenous and endogenous coordinates inside $x^f$.  For the smallest NK example,
where the states are only the exogenous demand and monetary-policy shocks,
this is not a serious modeling error.  For Rotemberg NK, SGU, EZ, and
NAWM-scale DSGE models, the state vector contains deterministic or
endogenous/predetermined components.  Treating those models as if every state
coordinate were an exchangeable noisy Gaussian coordinate is a structural
filtering bug, not a mere numerical tuning problem.

\section{Required Tests Before Any DSGE Nonlinear Filter Claim}
\label{sec:bf-structural-deterministic-required-tests}

No future BayesFilter report should claim that a nonlinear DSGE filter is
value-valid, gradient-valid, sampler-usable, converged, or production-ready in
the sense of Chapter~\ref{ch:bf-production-checklist} unless the following
structural tests are passed or explicitly waived with written justification:
\begin{enumerate}
  \item \textbf{Metadata test:} the model exposes exogenous, endogenous, and
    deterministic state blocks, and their dimensions match the shock-impact
    matrix and transition maps.
  \item \textbf{Constraint-support test:} propagated sigma points or particles
    satisfy deterministic identities to numerical tolerance.
  \item \textbf{Linear recovery test:} when the structural transition is
    linear, the structural nonlinear filter recovers the exact Kalman
    likelihood up to the declared approximation tolerance.
  \item \textbf{Degenerate-transition test:} zero-innovation endogenous rows
    remain deterministic except for explicitly declared numerical
    regularization.
  \item \textbf{Toy DSGE test:} a controlled model with
    \[
      m_t = \rho m_{t-1}+\sigma\varepsilon_t,\qquad
      k_t=f(k_{t-1},m_{t-1},m_t),
    \]
    compares structural propagation against a dense quadrature or
    high-particle reference.
  \item \textbf{Case-study test:} Rotemberg NK, SGU, and any EZ/NAWM target
    must document which state coordinates are shock-driven, which are
    deterministic conditional on the shock path, and how the filter honors
    that timing.
\end{enumerate}

\section{Common Misunderstandings}
\label{sec:bf-structural-common-misunderstandings}

The most common reader errors in this area are:
\begin{enumerate}[label=(\roman*)]
  \item a deterministic endogenous state need not have zero one-step predictive
    variance; it can inherit randomness through the declared stochastic block;
  \item a singular or rank-deficient transition covariance is not, by itself, a
    modeling error in the exact linear-Gaussian case;
  \item numerical regularization is not the same thing as declaring a new
    stochastic state coordinate;
  \item a full-state nonlinear approximation is not automatically forbidden, but
    it must be labeled as an approximation if it enlarges the stochastic space
    beyond the structural transition contract.
\end{enumerate}


The rule is simple:
\begin{center}
\fbox{\parbox{0.88\linewidth}{
For nonlinear DSGE filtering, the stochastic integration variables are not
automatically the entries of the state vector.  They are the variables declared
stochastic by the structural transition.  Deterministic endogenous states must
be computed, not noised.
}}
\end{center}

If an implementation violates this rule, it must be labeled as an artificial
approximation and justified against a reference.  Otherwise, the filter is
solving the wrong state-space model.  This policy should be treated as a release gate for any BayesFilter DSGE
backend intended for HMC, industrial forecasting, or peer-reviewed empirical
work.  Failing one of these tests does not automatically make a backend useless,
but it does mean the strongest claim labels from
Chapter~\ref{ch:bf-production-checklist} are not yet justified.
