\chapter{Particle-Filter Foundations}
\label{ch:bf-particle-filters}

This chapter develops the particle-filter baseline for the differentiable
particle-filter program.  The goal is not yet to make the filter
differentiable.  The goal is to establish, in BayesFilter notation, the
nonlinear filtering recursion, the empirical particle approximation, the
bootstrap particle-filter likelihood estimator, and the exact point at which the
later differentiable program departs from the classical construction.

Particle methods matter because the nonlinear filtering problem is generally not
closed under finite-dimensional moments.  When the predictive law or filtering
law is materially non-Gaussian, a Gaussian closure such as the EKF or UKF may be
a useful approximation but no longer defines the exact filtering object.  The
particle filter instead approximates the filtering law directly as a weighted
random measure.  This makes it the natural baseline for any later attempt to
construct particle-flow, resampling-relaxed, or differentiable particle
methods.

\section{Nonlinear state-space model and filtering recursion}
\label{sec:bf-pf-ssm-recursion}

Let $x_t \in \mathbb{R}^{n_x}$ denote the latent state and
$y_t \in \mathbb{R}^{n_y}$ the observation at time $t$.  Fix a parameter vector
$\theta \in \Theta \subset \mathbb{R}^{d_\theta}$.  The nonlinear state-space
model is specified by a transition density and an observation density,
\begin{align}
\label{eq:bf-pf-transition-density}
  x_t \mid x_{t-1} &\sim p_\theta(x_t \mid x_{t-1}), \\
\label{eq:bf-pf-observation-density}
  y_t \mid x_t &\sim p_\theta(y_t \mid x_t).
\end{align}
Write $y_{1:t}=(y_1,\ldots,y_t)$ for the observation history.  The predictive
law and the filtering law satisfy the standard Bayesian recursion
\begin{align}
\label{eq:bf-pf-predictive-law}
  p_\theta(x_t \mid y_{1:t-1})
  &= \int p_\theta(x_t \mid x_{t-1})
     p_\theta(x_{t-1} \mid y_{1:t-1})\,dx_{t-1}, \\
\label{eq:bf-pf-filtering-law}
  p_\theta(x_t \mid y_{1:t})
  &= \frac{p_\theta(y_t \mid x_t)
           p_\theta(x_t \mid y_{1:t-1})}
          {p_\theta(y_t \mid y_{1:t-1})},
\end{align}
where the one-step predictive density of the observation is
\begin{equation}
\label{eq:bf-pf-predictive-observation}
  p_\theta(y_t \mid y_{1:t-1})
  = \int p_\theta(y_t \mid x_t)
      p_\theta(x_t \mid y_{1:t-1})\,dx_t.
\end{equation}
The marginal likelihood then factorizes as
\begin{equation}
\label{eq:bf-pf-marginal-factorization}
  p_\theta(y_{1:T})
  = \prod_{t=1}^T p_\theta(y_t \mid y_{1:t-1}).
\end{equation}

Equations \eqref{eq:bf-pf-predictive-law}--\eqref{eq:bf-pf-marginal-factorization}
are exact.  The difficulty is computational rather than conceptual: outside
special classes such as the linear-Gaussian case, the integrals are rarely
available in closed form.  Particle filters replace these exact distributions by
random empirical approximations whose quality improves with particle count under
appropriate assumptions.

\section{Empirical filtering measures}
\label{sec:bf-pf-empirical-measures}

A particle filter represents the filtering law by a weighted empirical measure.
At time $t$, let
$\{(x_t^{(i)},w_t^{(i)})\}_{i=1}^N$ be particles and normalized weights, with
$w_t^{(i)}\ge 0$ and $\sum_{i=1}^N w_t^{(i)}=1$.  The empirical filtering
measure is
\begin{equation}
\label{eq:bf-pf-empirical-measure}
  \hat\pi_t^N(dx)
  = \sum_{i=1}^N w_t^{(i)}\,\delta_{x_t^{(i)}}(dx),
\end{equation}
where $\delta_z$ is the Dirac point mass at $z$.  The approximation of a test
function $\varphi$ under the filtering law is then
\begin{equation}
\label{eq:bf-pf-test-function-approx}
  \int \varphi(x)\,\hat\pi_t^N(dx)
  = \sum_{i=1}^N w_t^{(i)}\varphi(x_t^{(i)}).
\end{equation}
Thus the filter approximates the entire posterior law by a finite atomic
measure, not only its first two moments.

This is the main structural distinction between particle methods and the
Gaussian-approximation filters of the previous chapters.  The particle filter is
not built by closing the recursion in a finite-dimensional moment family.
Instead, it propagates a random discrete approximation to the law itself.

\section{Sequential importance sampling}
\label{sec:bf-pf-sis}

To derive the basic particle recursion, introduce a proposal density
$q_\theta(x_t \mid x_{0:t-1}, y_{1:t})$ from which particles can be sampled.
Suppose a full trajectory proposal factors as
\begin{equation}
\label{eq:bf-pf-trajectory-proposal}
  q_\theta(x_{0:t} \mid y_{1:t})
  = q_\theta(x_0 \mid y_1)
    \prod_{s=1}^t q_\theta(x_s \mid x_{0:s-1}, y_{1:s}).
\end{equation}
The target filtering distribution over trajectories is proportional to
\begin{equation}
\label{eq:bf-pf-trajectory-target}
  p_\theta(x_{0:t}, y_{1:t})
  = p_\theta(x_0)
    \prod_{s=1}^t p_\theta(x_s \mid x_{s-1})
    p_\theta(y_s \mid x_s).
\end{equation}
The unnormalized importance weight is therefore
\begin{equation}
\label{eq:bf-pf-importance-weight}
  \tilde w_t(x_{0:t})
  = \frac{p_\theta(x_{0:t},y_{1:t})}
         {q_\theta(x_{0:t} \mid y_{1:t})}.
\end{equation}
Using the factorization in
\eqref{eq:bf-pf-trajectory-proposal}--\eqref{eq:bf-pf-trajectory-target}, one
obtains the recursive update
\begin{equation}
\label{eq:bf-pf-sis-recursion}
  \tilde w_t^{(i)}
  = \tilde w_{t-1}^{(i)}
    \frac{p_\theta(y_t \mid x_t^{(i)})
          p_\theta(x_t^{(i)} \mid x_{t-1}^{(i)})}
         {q_\theta(x_t^{(i)} \mid x_{0:t-1}^{(i)}, y_{1:t})}.
\end{equation}
Normalizing gives
\begin{equation}
\label{eq:bf-pf-normalized-weights}
  w_t^{(i)}
  = \frac{\tilde w_t^{(i)}}{\sum_{j=1}^N \tilde w_t^{(j)}}.
\end{equation}

Equation \eqref{eq:bf-pf-sis-recursion} is the generic sequential importance
sampling (SIS) update.  It already shows the two main mathematical stresses of
particle filtering:
\begin{enumerate}[label=(\roman*)]
  \item the proposal must be chosen carefully to avoid explosive weight
    variability;
  \item even when the estimator is well defined, the normalized weights tend to
    concentrate over time, which leads to degeneracy.
\end{enumerate}

\section{Sequential importance resampling and the bootstrap filter}
\label{sec:bf-pf-sir-bootstrap}

Sequential importance sampling alone is typically unstable over long horizons,
because the weights become increasingly uneven.  Sequential importance
resampling (SIR) introduces a resampling step to refresh the cloud when the
weight distribution becomes too concentrated.

The most classical special case is the bootstrap particle filter, which uses the
transition prior itself as the proposal:
\begin{equation}
\label{eq:bf-pf-bootstrap-proposal}
  q_\theta(x_t \mid x_{0:t-1}, y_{1:t})
  = p_\theta(x_t \mid x_{t-1}).
\end{equation}
Substituting \eqref{eq:bf-pf-bootstrap-proposal} into
\eqref{eq:bf-pf-sis-recursion} gives the simplified incremental weight
\begin{equation}
\label{eq:bf-pf-bootstrap-weight}
  \tilde w_t^{(i)}
  \propto w_{t-1}^{(i)} p_\theta(y_t \mid x_t^{(i)}).
\end{equation}
The bootstrap recursion at time $t$ can therefore be written as:
\begin{enumerate}[label=\textbf{Step \arabic*:}]
  \item \textbf{Propagate:}
    sample
    $x_t^{(i)} \sim p_\theta(x_t \mid x_{t-1}^{(a_{t-1}^{(i)})})$.
  \item \textbf{Weight:}
    compute
    $\tilde w_t^{(i)} = p_\theta(y_t \mid x_t^{(i)})$
    and normalize the weights.
  \item \textbf{Resample:}
    draw ancestor indices using the normalized weights and reset weights to
    $1/N$.
\end{enumerate}
This yields the standard bootstrap/SIR particle filter of
\citet{gordon1993novel}, elaborated in the SMC text of
\citet{doucet2001sequential}.

\section{The bootstrap particle-filter likelihood estimator}
\label{sec:bf-pf-likelihood-estimator}

The bootstrap particle filter supplies a natural estimator of the factors in the
marginal-likelihood decomposition
\eqref{eq:bf-pf-marginal-factorization}.  Define the average incremental weight
at time $t$ by
\begin{equation}
\label{eq:bf-pf-average-weight}
  Z_t^N = \frac{1}{N}\sum_{i=1}^N \tilde w_t^{(i)}.
\end{equation}
The corresponding likelihood estimator is
\begin{equation}
\label{eq:bf-pf-bootstrap-likelihood-estimator}
  \hat p_\theta^N(y_{1:T}) = \prod_{t=1}^T Z_t^N.
\end{equation}
The estimator-status result BayesFilter needs at this stage is the following.

\begin{proposition}[Bootstrap particle-filter likelihood-estimator status]
\label{prop:bf-pf-bootstrap-likelihood-status}
Under the standard bootstrap-particle-filter assumptions, the estimator
\eqref{eq:bf-pf-bootstrap-likelihood-estimator} is an unbiased estimator of the
marginal likelihood:
\[
  \mathbb{E}\bigl[\hat p_\theta^N(y_{1:T})\bigr] = p_\theta(y_{1:T}).
\]
\end{proposition}

\begin{proof}[Proof sketch]
Conditional on the resampled cloud at time $t-1$, the bootstrap proposal draws
particles from the transition prior.  Therefore
\[
  \mathbb{E}\left[Z_t^N \mid x_{t-1}^{(1:N)}\right]
  = \frac{1}{N}\sum_{i=1}^N
    \int p_\theta(y_t \mid x_t)
         p_\theta(x_t \mid x_{t-1}^{(i)})\,dx_t.
\]
The integral equals the one-step predictive observation density conditional on
$x_{t-1}^{(i)}$.  Iterating this argument through the resampling step yields the
usual telescoping expectation identity for the product
$\prod_{t=1}^T Z_t^N$, giving the likelihood
$p_\theta(y_{1:T})$; see \citet{doucet2001sequential} and
\citet{andrieu2010particle} for the formal particle-system argument.
\end{proof}

This proposition is mathematically central for the later DPF program.  It tells
us that the classical bootstrap particle filter supplies an unbiased estimator
of the marginal likelihood, even though it does not supply a pathwise differentiable
likelihood map.  Later differentiable constructions must therefore be compared
against this baseline carefully: a differentiable surrogate may be smoother to
optimize or differentiate, but it need not preserve this exact estimator
status.

\section{Degeneracy and effective sample size}
\label{sec:bf-pf-degeneracy}

The main numerical weakness of the particle filter is weight degeneracy.  As the
observation sequence grows, the normalized weights frequently concentrate on a
small subset of particles.  A standard summary is the effective sample size
(ESS),
\begin{equation}
\label{eq:bf-pf-ess}
  \mathrm{ESS}_t = \frac{1}{\sum_{i=1}^N (w_t^{(i)})^2}.
\end{equation}
The extreme cases are informative:
\begin{itemize}
  \item if all weights are equal, then $\mathrm{ESS}_t=N$;
  \item if one particle carries almost all the mass, then
    $\mathrm{ESS}_t \approx 1$.
\end{itemize}
When ESS collapses, the atomic approximation
\eqref{eq:bf-pf-empirical-measure} still exists, but its practical variance may
become extremely large.  Resampling alleviates this by discarding low-weight
particles and duplicating high-weight particles, but resampling does not remove
the deeper dimensionality problem.  The classical literature shows that in many
high-dimensional settings the number of particles needed to stabilize the weight
distribution can grow very rapidly with the effective dimension of the
observation problem \citep{bengtsson2008curse,snyder2008obstacles}.

This is the exact point where the later chapters enter.  Particle-flow methods,
PF-PF corrections, differentiable resampling, OT relaxations, and learned OT
operators are all responses to some combination of:
\begin{enumerate}[label=(\roman*)]
  \item degeneracy under standard resampling;
  \item lack of differentiability of the resampling step;
  \item computational burden of high-quality resampling or transport maps;
  \item mismatch between a useful learning objective and a valid HMC target.
\end{enumerate}
A mathematically clean DPF monograph must therefore begin here: by making the
classical particle-filter object precise before modifying it.

\section{What this chapter does and does not claim}
\label{sec:bf-pf-chapter-boundary}

This chapter establishes the exact nonlinear filtering recursion and the
classical bootstrap particle-filter baseline.  It does \emph{not} yet claim any
of the following:
\begin{itemize}
  \item a pathwise differentiable likelihood map;
  \item a particle-flow approximation;
  \item a proposal-corrected flow filter;
  \item a differentiable resampling rule;
  \item an HMC-ready particle backend.
\end{itemize}
Those belong to the later chapters.  The role of the present chapter is more
basic and more important: it fixes the reference object against which every
later DPF construction must be interpreted.
