\chapter{Learned Transport Operators and Implementation Mathematics}
\label{ch:bf-dpf-learned-ot}

The previous chapter developed differentiable resampling up to the level of
entropic optimal transport, finite Sinkhorn iteration, and barycentric
projection.  This chapter studies a further approximation layer: learned or
amortized transport operators trained to imitate a previously defined transport
map.  The central point is deliberately conservative.  A learned operator is not
a speed trick that automatically preserves target correctness.  It is a student
map trained against a teacher object, and the teacher object may already be a
relaxed or finite-solver approximation.

The source spine is correspondingly narrow.  The entropy-regularized
particle-filter teacher comes from the differentiable PF construction of
\citet{corenflos2021differentiable}; Sinkhorn and computational OT background
comes from \citet{cuturi2013sinkhorn} and
\citet{peyre2019computational}; and the architecture discussion uses
permutation-aware set operators such as DeepSets and Set Transformer
\citep{zaheer2017deep,lee2019set}.  These sources identify plausible teacher
and architecture families.  They do not prove posterior preservation, HMC
target correctness, out-of-distribution reliability, or banking suitability for
a learned map.

\section{Objects and conventions}
\label{sec:bf-learned-ot-objects}

Let $X=(x_1,\ldots,x_N)^\top\in\mathbb{R}^{N\times d_x}$ be a particle cloud
and let $w\in\Delta_N$ be normalized weights.  Let
$u=(1/N,\ldots,1/N)$ be the equal-weight target marginal.  The cost matrix is
$C(X,Z)\in\mathbb{R}^{N\times M}$ for source cloud $X$ and target support
$Z=(z_1,\ldots,z_M)^\top$.  In the square resampling case used below,
$M=N$ and the barycentric output is an equal-weight cloud
$\tilde X\in\mathbb{R}^{N\times d_x}$.

\begin{longtable}{@{}p{0.18\textwidth} p{0.37\textwidth} p{0.36\textwidth}@{}}
\caption{Learned-OT object inventory.}
\label{tab:bf-learned-ot-object-inventory}\\
\toprule
Object & Definition & Audit question \\
\midrule
\endfirsthead
\toprule
Object & Definition & Audit question \\
\midrule
\endhead
Categorical baseline &
Realized ancestor resampling from weights $w$. &
Is the learned map being compared with categorical resampling, or only with an
OT teacher? \\
Unregularized OT teacher &
Optimizer $\Pi_0^\star$ of a transport objective with marginals $(w,u)$. &
Is the teacher the unregularized coupling, or is this only a limiting reference? \\
EOT teacher &
Optimizer $\Pi_\varepsilon^\star$ of an entropy-regularized OT objective. &
Is $\varepsilon$ fixed and recorded, and is the entropy convention stated? \\
Finite Sinkhorn teacher &
Finite-iteration coupling $\Pi_{\varepsilon,K}$ computed by a declared solver. &
Is the student trained on a numerical teacher rather than the exact EOT
optimizer? \\
Barycentric teacher output &
Equal-weight cloud $B(\Pi) \in \mathbb{R}^{N\times d_x}$ derived from a
coupling. &
Is the student imitating the coupling, the barycentric cloud, or a scalar
summary? \\
Learned student map &
Parametric map $T_\psi$ from weighted particle data and settings to an output
cloud or coupling. &
Are inputs, outputs, architecture class, particle count, dimension, and
$\varepsilon$ dependence declared? \\
Training distribution &
Distribution $\mathcal D$ over clouds, weights, model regimes, and solver
settings used to generate supervision. &
Is every learned-map claim restricted to the support or stress envelope of
$\mathcal D$? \\
Loss function &
Empirical objective used to fit the student to the teacher. &
Does the loss measure the object later claimed, or only a proxy such as map
MSE? \\
Permutation action &
Permutation matrix $P$ acting on particle rows and weights. &
Does the architecture respect row relabeling of an unordered particle cloud? \\
Residuals &
Teacher-object, solver, learning, distribution-shift, filtering, posterior, and
HMC value-gradient errors. &
Is each residual tied to a mathematical object and a stated ``does not prove''
limit? \\
\bottomrule
\end{longtable}

\section{Teacher variants}
\label{sec:bf-learned-ot-hierarchy}

There is no single object called ``the OT teacher'' unless the variant is named.
For a cost matrix $C$ and feasible set
\[
  \mathcal U(w,u)=
  \left\{\Pi\in\mathbb{R}_+^{N\times N}:
  \Pi\mathbf 1=w,\ \Pi^\top\mathbf 1=u\right\},
\]
the unregularized teacher is
\begin{equation}
\label{eq:bf-learned-ot-unregularized-teacher}
  \Pi_0^\star
  \in
  \arg\min_{\Pi\in\mathcal U(w,u)}
  \langle C(X), \Pi\rangle .
\end{equation}
This teacher is a transport projection, not categorical ancestor sampling.  The
EOT teacher for the entropy convention used in the preceding chapter is
\begin{equation}
\label{eq:bf-learned-ot-eot-teacher}
  \Pi_\varepsilon^\star
  \in
  \arg\min_{\Pi\in\mathcal U(w,u)}
  \left\{
    \langle C(X), \Pi\rangle
    + \varepsilon\sum_{i,j}\Pi_{ij}(\log \Pi_{ij}-1)
  \right\},
  \qquad \varepsilon>0 .
\end{equation}
The finite numerical teacher is the output of a declared Sinkhorn procedure:
\begin{equation}
\label{eq:bf-learned-ot-finite-teacher}
  \Pi_{\varepsilon,K}=\operatorname{Sinkhorn}_K(C(X),w,u,\varepsilon),
\end{equation}
where $K$ records the iteration budget and the implementation also fixes
stabilization, tolerances, and stopping policy.  In many amortized designs the
student is trained on $\Pi_{\varepsilon,K}$ or on its barycentric projection,
not on $\Pi_\varepsilon^\star$ and not on $\Pi_0^\star$.

For a coupling $\Pi$, define the equal-weight barycentric output, for
$j\in\{1,\ldots,N\}$, by
\begin{equation}
\label{eq:bf-learned-ot-barycentric-teacher}
  B(\Pi)_j
  =
  \frac{\sum_{i=1}^N \Pi_{ij} x_i}{\sum_{i=1}^N \Pi_{ij}},
\end{equation}
whenever the denominator is positive.  For the target marginal $u_j=1/N$, this
is $B(\Pi)_j=N\sum_i \Pi_{ij}x_i$.  A training set must state whether its
teacher output is $B(\Pi_0^\star)$, $B(\Pi_\varepsilon^\star)$,
$B(\Pi_{\varepsilon,K})$, the coupling itself, or a lower-dimensional summary.

\begin{longtable}{@{}p{0.18\textwidth} p{0.30\textwidth} p{0.24\textwidth} p{0.20\textwidth}@{}}
\caption{Teacher variants and what they do not prove.}
\label{tab:bf-learned-ot-teacher-variants}\\
\toprule
Teacher variant & Object actually used & Direct status & Does not prove \\
\midrule
\endfirsthead
\toprule
Teacher variant & Object actually used & Direct status & Does not prove \\
\midrule
\endhead
Categorical baseline &
Ancestor draw from $\operatorname{Categorical}(w)$. &
Classical resampling law, discontinuous pathwise. &
Pathwise differentiability or learned-map correctness. \\
Unregularized OT &
$\Pi_0^\star$ or $B(\Pi_0^\star)$. &
Exact for the chosen OT projection. &
Equivalence to categorical resampling or EOT. \\
EOT &
$\Pi_\varepsilon^\star$ or $B(\Pi_\varepsilon^\star)$. &
Exact for the regularized objective. &
Small $\varepsilon$ stability, finite-solver accuracy, or posterior invariance. \\
Finite Sinkhorn &
$\Pi_{\varepsilon,K}$ or $B(\Pi_{\varepsilon,K})$. &
Numerical approximation to the EOT teacher. &
Convergence to EOT, absence of stabilization artifacts, or HMC validity. \\
Learned student supervision &
Samples from whichever teacher was computed. &
Empirical approximation problem on $\mathcal D$. &
Out-of-distribution accuracy, posterior preservation, or bank-facing evidence. \\
\bottomrule
\end{longtable}

\section{Student map and training objective}
\label{sec:bf-learned-ot-teacher}

A learned or amortized transport operator is a parametric map.  In the
barycentric-output version, write
\begin{equation}
\label{eq:bf-learned-ot-student-map}
  \tilde X_\psi
  =
  T_\psi(X,w;\varepsilon,N,d_x,a),
\end{equation}
where $\psi$ are trainable parameters and $a$ records architecture choices such
as shared-row networks, pooling, attention blocks, depth, width, and positional
or feature encodings.  The dependence on $N$, $d_x$, and $\varepsilon$ is not a
software detail.  It says which mathematical family of maps is being fit.

If the teacher output is
\[
  Y_\tau(X,w,\varepsilon)=B(\Pi_\tau(X,w,\varepsilon)),
\]
where $\tau\in\{0,\varepsilon,(\varepsilon,K)\}$ names the teacher variant, a
supervised objective has the form
\begin{equation}
\label{eq:bf-learned-ot-training-objective}
  \psi^\star
  \in
  \arg\min_\psi
  \mathbb E_{(X,w,\varepsilon,m)\sim \mathcal D}
  \left[
    \mathcal L\{T_\psi(X,w;\varepsilon,N,d_x,a),Y_\tau(X,w,\varepsilon)\}
  \right],
\end{equation}
where $m$ denotes the state-space model or regime that generated the cloud.  In
practice the expectation is replaced by an empirical average:
\begin{equation}
\label{eq:bf-learned-ot-empirical-training-objective}
  \widehat J(\psi)
  =
  \frac{1}{S}
  \sum_{s=1}^S
  \mathcal L\{T_\psi(X^{(s)},w^{(s)};\varepsilon^{(s)},N^{(s)},d_x^{(s)},a),
  Y_\tau(X^{(s)},w^{(s)},\varepsilon^{(s)})\}.
\end{equation}

Equations~\eqref{eq:bf-learned-ot-training-objective} and
\eqref{eq:bf-learned-ot-empirical-training-objective} prove only what they
define: an empirical teacher-student fitting problem.  They do not prove small
filtering error, posterior error, HMC target correctness, or validity outside
the training distribution.  Those require additional diagnostics tied to the
filtering recursion and posterior target.

\section{Permutation symmetry}
\label{sec:bf-learned-ot-symmetry}

Particle labels are arbitrary.  Let $P\in\mathbb{R}^{N\times N}$ be a
permutation matrix.  For output clouds whose rows are tied to the same particle
ordering convention, the natural requirement is permutation equivariance:
\begin{equation}
\label{eq:bf-learned-ot-equivariance}
  T_\psi(PX,Pw;\varepsilon,N,d_x,a)
  =
  P\,T_\psi(X,w;\varepsilon,N,d_x,a).
\end{equation}
For scalar summaries $s_\psi(X,w)$, the corresponding requirement is
permutation invariance:
\begin{equation}
\label{eq:bf-learned-ot-invariance}
  s_\psi(PX,Pw)=s_\psi(X,w).
\end{equation}
DeepSets-style shared row maps and global aggregation
\citep{zaheer2017deep}, and attention-based set operators
\citep{lee2019set}, are relevant because they can encode such symmetries.

Symmetry is necessary but not target-sufficient.  A perfectly equivariant
student can still approximate the wrong teacher, extrapolate outside
$\mathcal D$, collapse the output cloud, or produce a value-gradient path that
does not match the scalar target used by HMC.  Conversely, symmetry violation is
already a mathematical failure: the output measure depends on arbitrary input
ordering rather than on the empirical measure represented by the cloud.

\section{Approximation residuals and target shift}
\label{sec:bf-learned-ot-residuals}

Residuals must be named by object.  A small residual at one layer does not
automatically bound a later residual.  The map-level residual against the
teacher $\tau$ is
\begin{equation}
\label{eq:bf-learned-ot-residual}
  R_{\psi,\tau}(X,w,\varepsilon)
  =
  T_\psi(X,w;\varepsilon,N,d_x,a)-Y_\tau(X,w,\varepsilon).
\end{equation}
If the teacher is finite Sinkhorn, there is also a teacher-object residual:
\begin{equation}
\label{eq:bf-learned-ot-teacher-object-residual}
  E_{\mathrm{teach}}(X,w,\varepsilon,K)
  =
  B(\Pi_{\varepsilon,K})-B(\Pi_\varepsilon^\star).
\end{equation}
A filtering-summary residual can be written for a test function $\varphi$ as
\begin{equation}
\label{eq:bf-learned-ot-filtering-summary-residual}
  E_\varphi
  =
  \frac{1}{N}\sum_{j=1}^N \varphi\{T_\psi(X,w)_j\}
  -
  \frac{1}{N}\sum_{j=1}^N \varphi\{Y_\tau(X,w)_j\}.
\end{equation}
An HMC value-gradient residual for a learned scalar $\ell_\psi$ and a teacher
scalar $\ell_\tau$ is different again:
\begin{equation}
\label{eq:bf-learned-ot-hmc-gradient-residual}
  E_{\nabla}
  =
  \nabla_\theta \ell_\psi(\theta)-\nabla_\theta \ell_\tau(\theta).
\end{equation}
This residual does not decide whether HMC is correct.  It measures discrepancy
between two targets.  HMC target correctness still requires the gradient used by
the integrator to be the derivative of the same scalar value used in the
Metropolis correction.

\begin{longtable}{@{}p{0.16\textwidth} p{0.25\textwidth} p{0.26\textwidth} p{0.24\textwidth}@{}}
\caption{Residual hierarchy for learned OT.}
\label{tab:bf-learned-ot-residual-hierarchy}\\
\toprule
Residual layer & Direct object & Does not prove & Required next diagnostic \\
\midrule
\endfirsthead
\toprule
Residual layer & Direct object & Does not prove & Required next diagnostic \\
\midrule
\endhead
Teacher-object error &
$B(\Pi_{\varepsilon,K})-B(\Pi_\varepsilon^\star)$ or EOT versus unregularized OT. &
That the teacher matches categorical resampling or the exact posterior. &
Solver residuals, $\varepsilon$ sensitivity, and teacher-target comparison. \\
Finite-solver error &
Marginal residuals, cost gap, and stopping residuals of Sinkhorn. &
That the learned student is accurate or stable. &
Fixed-iteration sensitivity and stabilized versus unstabilized comparisons. \\
Supervised learning error &
$R_{\psi,\tau}$ on training and held-out clouds. &
Small filtering-summary error or posterior error. &
Stratified errors by weight degeneracy, dimension, particle count, and
$\varepsilon$. \\
Distribution shift &
Difference between deployment clouds and $\mathcal D$. &
That held-out in-distribution residuals transfer. &
Stress sets for sharp weights, high dimension, particle-count shift, and model
regions. \\
Filtering-summary error &
Test-function errors such as $E_\varphi$. &
Global target equivalence. &
Multiple nonlinear test functions and full filtering likelihood comparisons. \\
Posterior error &
Parameter posterior difference under learned and teacher filters. &
HMC value-gradient consistency or production suitability. &
Posterior comparisons against teacher OT and trusted references. \\
HMC value-gradient error &
Difference between gradients or scalar paths used by HMC. &
Correctness for either target unless the value-gradient contract is satisfied. &
Same-scalar value/gradient checks, finite differences, and accept/reject audit. \\
\bottomrule
\end{longtable}

\section{Training-distribution dependence}
\label{sec:bf-learned-ot-training-distribution}

The training distribution must be part of the mathematical claim.  Write
\[
  \mathcal D
  =
  \mathcal D(N,d_x,\varepsilon,m,\eta),
\]
where $m$ is the model class or simulation regime and $\eta$ records weight
geometry, particle degeneracy, missing-data pattern, and other cloud-generating
features.  A claim about $T_\psi$ is local to the support and stress envelope of
this distribution unless a separate generalization argument is given.

Important distribution coordinates include:
\begin{itemize}
  \item state dimension $d_x$ and latent-factor structure;
  \item particle count $N$ and any batching or padding convention;
  \item weight-simplex geometry, especially sharp or nearly collapsed weights;
  \item regularization range for $\varepsilon$ and Sinkhorn iteration budget
    $K$;
  \item structural-model region, including invalid or near-boundary parameter
    values;
  \item missing-data, mixed-frequency, and long-panel regimes;
  \item hardware or compiled-execution regime if the learned map is used inside
    repeated likelihood evaluations.
\end{itemize}

The phrase ``the learned map works'' is therefore not acceptable by itself.  A
defensible statement has the form: relative to teacher $\tau$, training
distribution $\mathcal D$, architecture class $a$, particle count and dimension
envelope, and recorded stress tests, the student residuals remain below stated
thresholds.  Even that statement remains a teacher-student statement until
filtering and posterior diagnostics are added.

\section{Failure regimes}
\label{sec:bf-learned-ot-failure-regimes}

\begin{longtable}{@{}p{0.18\textwidth} p{0.29\textwidth} p{0.24\textwidth} p{0.21\textwidth}@{}}
\caption{Failure regimes for learned OT.}
\label{tab:bf-learned-ot-failure-regimes}\\
\toprule
Regime & Why it is hard & Likely failure mode & Required diagnostic \\
\midrule
\endfirsthead
\toprule
Regime & Why it is hard & Likely failure mode & Required diagnostic \\
\midrule
\endhead
Sharp weights &
Mass concentrates on few particles. &
Student smooths or collapses the cloud when resampling matters most. &
Residuals stratified by ESS and maximum weight. \\
High dimension &
Transport geometry and network capacity change with $d_x$. &
Low-dimensional residuals do not transfer. &
Dimension ladder and coordinate-scaling tests. \\
Particle-count shift &
Deployment $N$ differs from training $N$. &
Architecture, normalization, or attention cost changes the learned map. &
Train/test ladders in $N$ with fixed model and fixed $\varepsilon$. \\
$\varepsilon$ shift &
Regularization controls both teacher smoothness and target drift. &
Student trained at one smoothness level is used at another. &
Sensitivity over $\varepsilon$ and teacher residual comparisons. \\
Structural-model regions &
Invalid, near-determinacy, or stiff parameter regions change cloud geometry. &
Student extrapolates where the filter is already fragile. &
DSGE/MacroFinance stress clouds, invalid-region policy, and failure logging. \\
Learned-map collapse &
Network maps diverse inputs to low-diversity outputs. &
Particle diversity and likelihood curvature are distorted. &
Output covariance, duplicate-count, and nonlinear test-function diagnostics. \\
Compiled repetition &
The map is used repeatedly inside a likelihood or HMC path. &
Tiny deterministic biases accumulate across long panels. &
Long-panel repeated-evaluation tests and seed/hardware reproducibility. \\
\bottomrule
\end{longtable}

\section{Implementation-facing mathematics}
\label{sec:bf-learned-ot-implementation-math}

Runtime gains are real engineering evidence, but they are not target evidence.
If $T_\psi$ replaces a Sinkhorn solve by a single forward pass, the cost can
fall substantially.  The reason is that the computation has changed from a
solver for a declared teacher to an evaluation of a fitted student.  A runtime
table must therefore sit next to teacher residuals, filtering-summary
diagnostics, and posterior comparisons.  It cannot replace them.

Deterministic evaluation is similarly double-edged.  A deterministic learned
map can simplify autodiff and repeated HMC likelihood evaluation compared with a
stochastic resampling path.  But determinism only makes the surrogate easier to
differentiate; it does not identify the posterior.  The HMC question remains:
what scalar value is being evaluated, and is the supplied gradient the
derivative of that same scalar?

Architecture choices also have mathematical status.  Shared-row maps, pooling,
attention blocks, equivariance constraints, fixed particle counts, padding, and
normalization layers define the class of transport maps available to the
student.  They are not software-only details.  When a learned OT result is
reported, the architecture is part of the approximation theorem that the work
does not yet have.

\section{Internal evidence ladder for bank-facing use}
\label{sec:bf-learned-ot-banking-evidence}

\begin{longtable}{@{}p{0.18\textwidth} p{0.27\textwidth} p{0.27\textwidth} p{0.19\textwidth}@{}}
\caption{Internal evidence ladder for learned-OT use in bank-facing filtering.}
\label{tab:bf-learned-ot-banking-evidence}\\
\toprule
Gate & Evidence required & What the evidence supports & What it does not support \\
\midrule
\endfirsthead
\toprule
Gate & Evidence required & What the evidence supports & What it does not support \\
\midrule
\endhead
Teacher residual tests &
Marginal residuals, cost gaps, $\varepsilon$ sensitivity, and finite-Sinkhorn
stability. &
The teacher computation is numerically interpretable. &
Learned-map accuracy or posterior preservation. \\
Student residual tests &
Train/held-out residuals by $N$, $d_x$, ESS, $\varepsilon$, and model regime. &
The student imitates the chosen teacher on the tested envelope. &
Global generalization or HMC target correctness. \\
Filtering stress tests &
Test-function errors, likelihood comparisons, cloud-diversity diagnostics, and
failure logs. &
The learned map is not visibly breaking the filtering summaries tested. &
Posterior equivalence or bank-facing deployment. \\
Posterior comparison &
Teacher versus student posterior summaries on named datasets and model classes. &
Credible research evidence for the tested posterior quantities. &
Production readiness or untested structural regimes. \\
Gradient and HMC checks &
Same-scalar value/gradient checks, finite differences, compiled repetition, and
accept/reject path audit. &
HMC coherence for the learned scalar if the target contract passes. &
Correctness for the original posterior unless corrected. \\
Governance limitations &
Versioned model, data, architecture, weights, diagnostics, monitoring, and
rollback plan. &
A controlled internal research or governance package after review. &
Automatic approval from speed, smoothness, or a small residual table. \\
\bottomrule
\end{longtable}

This internal ladder intentionally prevents a common overclaim.  It is not a
substitute for a bank's formal model-risk or production-governance framework.
A learned map can be fast, smooth, equivariant, and close to a finite Sinkhorn
teacher on a held-out set, while still lacking evidence for posterior
preservation under nonlinear DSGE or MacroFinance targets.  Bank-facing
credibility requires the later gates, not only the early ones, and formal use
would require a separate institution-specific governance process.

\section{Why this chapter stops short of an HMC verdict}
\label{sec:bf-learned-ot-no-hmc-verdict}

This chapter identifies learned or amortized OT as a layered surrogate above an
OT or finite-Sinkhorn teacher.  That is enough to prepare the HMC target
chapter, but it does not settle the HMC question.  The unresolved HMC question
requires a joint analysis of:
\begin{itemize}
  \item the baseline particle-filter or structural filtering target;
  \item the OT relaxation layer;
  \item the finite-solver layer, if the teacher is numerical;
  \item the learned-student approximation layer;
  \item the scalar value and gradient path supplied to HMC;
  \item the correction mechanism, if any.
\end{itemize}
If HMC uses the learned scalar and its matching gradient, the result may be
coherent for the learned target.  It is not thereby coherent for the original
categorical or structural posterior unless a correction or posterior-error
argument is supplied.

\section{What this chapter does and does not claim}
\label{sec:bf-learned-ot-boundary}

This chapter claims that learned OT must be documented as a teacher-student
surrogate with explicit residual and extrapolation risks.  It does not claim:
\begin{itemize}
  \item that learned OT is equivalent to the OT teacher;
  \item that an OT teacher is equivalent to categorical resampling;
  \item that a finite Sinkhorn teacher is the exact EOT optimizer;
  \item that a small map-level residual implies small filtering or posterior
    error;
  \item that permutation equivariance implies target correctness;
  \item that runtime improvement is evidence of posterior preservation;
  \item that learned OT is justified as a final HMC target for nonlinear DSGE or
    MacroFinance models;
  \item that learned OT is ready for bank-facing or production-governance use.
\end{itemize}
The chapter's role is narrower and mathematically prior: it identifies the
teacher actually being approximated, the student map being trained, the
distribution on which the claim is local, and the residual ladder that must be
closed before stronger target or banking language would be defensible.

\FloatBarrier
