\chapter{DPF Literature-to-Debugging Verification Contract}
\label{ch:bf-dpf-debug-crosswalk}

The preceding DPF chapters separate the mathematical layers: classical
particle filtering, particle-flow transport, proposal correction,
differentiable resampling, OT/Sinkhorn relaxation, learned transport, and HMC
target interpretation.  This chapter turns that separation into a verification
contract.  Its purpose is not to validate any implementation by prose.  Its
purpose is to tell a coding agent or reviewer which equation, controlled
example, implementation quantity, tolerance, and failure interpretation belongs
to each layer.

The contract is deliberately ordered.  Target and value-gradient checks come
before tuning.  Teacher-map diagnostics come before learned-map diagnostics.
Equation-local failures should be interpreted narrowly: a failed Sinkhorn
marginal test is evidence about a finite solver, not evidence that particle
flows are wrong; a learned-map residual is evidence about the student and its
training distribution, not evidence that the OT teacher is invalid.

\section{Diagnostic order}
\label{sec:bf-dpf-debug-order}

Use this order before changing code or tuning numerical parameters:
\begin{enumerate}[label=(\arabic*)]
  \item name the scalar object being estimated, relaxed, learned, or sampled;
  \item verify that the gradient path corresponds to that same scalar object
    whenever HMC or autodiff claims are made;
  \item test the classical particle-filter recursion and likelihood estimator
    on a controlled model;
  \item test the particle-flow endpoint and homotopy derivative before testing
    PF-PF proposal correction;
  \item test PF-PF weights and log-determinant accounting before interpreting
    resampling or OT behavior;
  \item test categorical, soft, EOT, Sinkhorn, and barycentric objects before
    testing a learned student map;
  \item only after those checks pass, run posterior comparisons and sampler
    diagnostics.
\end{enumerate}
This order prevents a common error: making a numerically smooth path smoother
without first proving that it is the intended path.

\section{Equation-to-test matrix}
\label{sec:bf-dpf-debug-equation-test-matrix}

\begin{center}
\scriptsize
\begin{longtable}{@{}p{0.13\textwidth} p{0.16\textwidth} p{0.17\textwidth} p{0.17\textwidth} p{0.15\textwidth} p{0.14\textwidth}@{}}
\caption{Equation-indexed verification matrix for the DPF block.}
\label{tab:bf-dpf-debug-equation-test-matrix}\\
\toprule
Layer & Equation or claim & Controlled example & Implementation quantity & Pass signal & Narrow failure interpretation \\
\midrule
\endfirsthead
\caption[]{Equation-indexed verification matrix for the DPF block, continued.}\\
\toprule
Layer & Equation or claim & Controlled example & Implementation quantity & Pass signal & Narrow failure interpretation \\
\midrule
\endhead
\bottomrule
\endlastfoot
Filtering recursion &
\eqref{eq:bf-pf-predictive-law}, \eqref{eq:bf-pf-filtering-law} &
linear-Gaussian state-space model with Kalman reference &
predict/update law, normalized weights, filtered mean/covariance &
moments match reference within declared Monte Carlo error &
recursion, normalization, or model-interface bug; not an OT or learned-map claim \\
Bootstrap likelihood estimator &
\eqref{eq:bf-pf-average-weight}, \eqref{eq:bf-pf-bootstrap-likelihood-estimator} &
same linear-Gaussian model; fixed seeds and increasing $N$ &
per-period average weights and log likelihood estimator &
Monte Carlo mean approaches Kalman likelihood; variance decreases with $N$ &
particle proposal or weight bug; not evidence about differentiability \\
Homotopy derivative &
\eqref{eq:bf-pff-log-homotopy}, \eqref{eq:bf-pff-log-homotopy-derivative} &
scalar nonlinear observation with analytic log-likelihood derivative &
finite difference in $\lambda$ versus analytic derivative &
central difference and analytic derivative agree under step refinement &
homotopy algebra or likelihood derivative issue; not yet a PF-PF correction issue \\
EDH endpoint &
\eqref{eq:bf-pff-edh-velocity}, \eqref{eq:bf-pff-kalman-endpoint} &
linear-Gaussian recovery with known posterior mean/covariance &
flow endpoint particles, mean, covariance, and endpoint residual &
endpoint moments match Kalman update within flow and Monte Carlo tolerances &
flow ODE or Gaussian-closure implementation issue; not HMC target validity \\
LEDH local endpoint &
\eqref{eq:bf-pff-local-jacobian}, \eqref{eq:bf-pff-ledh-ode} &
synthetic observation with known local Jacobian variation &
particle-local Jacobians, local precision, and endpoint continuity &
local quantities are finite and stable under small perturbations &
local linearization fragility; not proof that global EDH is wrong \\
PF-PF weight &
\eqref{eq:bf-pfpf-preflo-density}, \eqref{eq:bf-pfpf-postflow-density}, \eqref{eq:bf-pfpf-weight} &
synthetic affine flow with known determinant &
proposal density, target density, corrected log weight &
manual density ratio and implemented weight agree &
proposal-density or correction algebra bug; not a resampling diagnosis \\
Log-determinant ODE &
\eqref{eq:bf-pfpf-jacobian-ode}, \eqref{eq:bf-pfpf-logdet-ode}, \eqref{eq:bf-pfpf-logdet-affine} &
affine flow $x_\lambda=A_\lambda x+b_\lambda$ &
integrated log determinant and finite-difference determinant &
sign and magnitude match under step-size refinement &
Jacobian sign, time direction, or integrator issue; not a likelihood theorem \\
Categorical discontinuity &
\eqref{eq:bf-dr-categorical-selection}, \eqref{eq:bf-dr-two-particle-discontinuity} &
two-particle resampling with weight crossing &
ancestor index path and realized resampled cloud &
jump occurs as expected; no false pathwise gradient is reported &
discrete resampling is being differentiated incorrectly; not proof soft resampling is valid \\
Soft-resampling bias &
\eqref{eq:bf-dr-soft-resampling}, \eqref{eq:bf-dr-soft-test-bias} &
two-particle nonlinear test function &
mean preservation and nonlinear test-function error &
affine tests behave as derived; nonlinear bias is visible when expected &
surrogate-law effect; not evidence of categorical-target preservation \\
EOT/Sinkhorn marginals &
\eqref{eq:bf-dr-eot-primal}, \eqref{eq:bf-dr-sinkhorn-marginals}, \eqref{eq:bf-dr-sinkhorn-residuals} &
small cost matrix with prescribed marginals &
row residual, column residual, cost, entropy, $\varepsilon$, iteration count &
residuals fall below tolerance and are stable under log-domain implementation &
finite-solver or stabilization issue; not proof of posterior invariance \\
Barycentric projection &
\eqref{eq:bf-dr-barycentric-map} &
small coupling with known column masses &
projected equal-weight support cloud and dimensions &
output has declared shape and column-mass normalization &
projection or shape bug; not a coupling-optimality result \\
Learned-map residual &
\eqref{eq:bf-learned-ot-student-map}, \eqref{eq:bf-learned-ot-residual}, \eqref{eq:bf-learned-ot-equivariance} &
permutation-equivariance test and held-out teacher clouds &
teacher/student residuals, permutation residual, training-distribution tags &
residuals stay below threshold on the declared envelope &
student or distribution-shift issue; not a teacher-map refutation \\
HMC value-gradient contract &
\eqref{eq:bf-dpf-hmc-contract} &
finite-difference check on a fixed scalar target and fixed random seed policy &
scalar value, autodiff gradient, finite-difference gradient, accept/reject value &
all paths refer to the same scalar and gradients agree within tolerance &
target-interface bug; do not tune HMC before resolving it \\
\end{longtable}
\end{center}

\section{Minimal controlled examples}
\label{sec:bf-dpf-debug-controlled-examples}

\begin{longtable}{@{}p{0.18\textwidth} p{0.24\textwidth} p{0.26\textwidth} p{0.22\textwidth}@{}}
\caption{Minimal controlled examples for DPF verification.}
\label{tab:bf-dpf-debug-controlled-examples}\\
\toprule
Example & Required inputs & Expected output & Does not prove \\
\midrule
\endfirsthead
\toprule
Example & Required inputs & Expected output & Does not prove \\
\midrule
\endhead
Linear-Gaussian recovery &
Known transition, observation, covariance matrices, Kalman likelihood, and
Kalman filtering moments. &
Particle recursion, EDH endpoint, and likelihood estimates approach the Kalman
reference under increasing particles or decreasing flow step size. &
Nonlinear DSGE validity or learned-map generalization. \\
Synthetic affine flow &
Invertible affine map, analytic determinant, and known proposal density. &
PF-PF density ratio, weight, and log determinant match hand calculation. &
Correctness for nonlinear flow fields. \\
Scalar nonlinear observation &
One-dimensional observation function with analytic derivatives and controlled
curvature. &
Homotopy derivative and local-linearization diagnostics show expected error as
curvature changes. &
High-dimensional stability. \\
Two-particle soft-resampling bias &
Two particles, weights crossing $1/2$, and nonlinear test function. &
Categorical path is discontinuous; soft path has the predicted surrogate bias. &
Acceptability of soft resampling as an HMC target. \\
Small Sinkhorn problem &
Positive cost matrix, weights, equal target marginal, fixed $\varepsilon$, and
known residual tolerance. &
Row/column residuals and stabilized objective behave consistently with the
solver budget. &
Posterior equivalence to categorical resampling. \\
Permutation-equivariance learned-map test &
Teacher clouds, learned map, and permutation matrices. &
Permuting input rows permutes output rows, and scalar summaries remain
invariant when required. &
Small teacher/student error or target preservation. \\
Finite-difference HMC gradient test &
Fixed scalar target, fixed random seed policy, finite-difference step ladder,
and autodiff gradient. &
Autodiff and finite-difference gradients agree for the scalar used by
accept/reject. &
Convergence of the Markov chain or correctness for another scalar target. \\
\bottomrule
\end{longtable}

\section{Diagnostic specification}
\label{sec:bf-dpf-debug-diagnostic-spec}

Every diagnostic result should be recorded with the following fields:
\begin{itemize}
  \item \textbf{Equation or claim}: label from the chapter, not only a method
    name.
  \item \textbf{Inputs}: model, particles, weights, random seed policy,
    $\varepsilon$, iteration budget, architecture, and scalar target as
    applicable.
  \item \textbf{Expected output}: scalar, vector, matrix, empirical measure,
    residual, or posterior summary.
  \item \textbf{Tolerance}: absolute and relative thresholds, Monte Carlo
    standard-error logic, or convergence slope.
  \item \textbf{Failure interpretation}: the narrow mathematical layer that the
    failure implicates.
  \item \textbf{Non-implication}: what the failure does not prove.
\end{itemize}

\begin{longtable}{@{}p{0.15\textwidth} p{0.21\textwidth} p{0.21\textwidth} p{0.19\textwidth} p{0.16\textwidth}@{}}
\caption{Diagnostic contract by layer.}
\label{tab:bf-dpf-debug-diagnostic-contract}\\
\toprule
Diagnostic & Inputs & Tolerance rule & Failure interpretation & Non-implication \\
\midrule
\endfirsthead
\toprule
Diagnostic & Inputs & Tolerance rule & Failure interpretation & Non-implication \\
\midrule
\endhead
Filtering recursion parity &
Kalman reference, fixed particles, fixed model. &
Monte Carlo error decreases with $N$; deterministic quantities match reference
when applicable. &
State-space interface, normalization, or weight update bug. &
Does not reject OT, learned OT, or HMC. \\
Flow endpoint parity &
Linear-Gaussian recovery, flow step ladder. &
Endpoint error decreases as flow integration is refined. &
Flow ODE, closure, or integration issue. &
Does not prove PF-PF correction is wrong. \\
PF-PF algebra parity &
Affine flow, analytic determinant, explicit proposal density. &
Log-weight difference below deterministic tolerance. &
Density-ratio, Jacobian, or sign error. &
Does not diagnose resampling. \\
Soft-resampling bias check &
Two-particle nonlinear test. &
Bias sign and magnitude match derived expression within arithmetic tolerance. &
Surrogate target is being confused with categorical target. &
Does not prove soft resampling is unusable. \\
Sinkhorn residual check &
Small cost matrix, fixed $\varepsilon$, solver budget. &
Row/column residuals below declared threshold and improve with budget. &
Finite-solver or stabilization issue. &
Does not prove EOT is a valid posterior target. \\
Learned residual check &
Teacher outputs, held-out clouds, distribution tags. &
Residuals reported by regime; thresholds tied to training envelope. &
Student approximation or extrapolation issue. &
Does not invalidate the teacher. \\
HMC value-gradient check &
Scalar value, autodiff path, finite-difference ladder. &
Gradient errors shrink under step refinement until numerical limits. &
Value-gradient mismatch or nondifferentiable scalar path. &
Does not prove chain convergence. \\
\bottomrule
\end{longtable}

\section{Source and implementation map}
\label{sec:bf-dpf-debug-source-implementation-map}

\begin{center}
\scriptsize
\begin{longtable}{@{}p{0.14\textwidth} p{0.18\textwidth} p{0.20\textwidth} p{0.18\textwidth} p{0.20\textwidth}@{}}
\caption{Source, chapter, equation, and implementation links.}
\label{tab:bf-dpf-debug-source-implementation-map}\\
\toprule
Layer & Chapter location & Equation labels & Source family & Implementation quantity \\
\midrule
\endfirsthead
\toprule
Layer & Chapter location & Equation labels & Source family & Implementation quantity \\
\midrule
\endhead
Classical PF &
Chapter~\ref{ch:bf-particle-filters}, Sections~\ref{sec:bf-pf-ssm-recursion}--\ref{sec:bf-pf-likelihood-estimator} &
\eqref{eq:bf-pf-filtering-law}, \eqref{eq:bf-pf-bootstrap-likelihood-estimator} &
SMC literature spine &
weights, ESS, likelihood estimator, empirical moments \\
Particle flow &
Chapter~\ref{ch:bf-dpf-particle-flow-foundations}, Sections~\ref{sec:bf-pff-homotopy}--\ref{sec:bf-pff-ledh} &
\eqref{eq:bf-pff-log-homotopy-derivative}, \eqref{eq:bf-pff-edh-ode}, \eqref{eq:bf-pff-ledh-ode} &
Daum--Huang, PF-PF flow literature &
flow field, endpoint residual, local Jacobians \\
PF-PF correction &
Chapter~\ref{ch:bf-dpf-pfpf}, Sections~\ref{sec:bf-pfpf-change-of-variables}--\ref{sec:bf-pfpf-logdet} &
\eqref{eq:bf-pfpf-weight}, \eqref{eq:bf-pfpf-logdet-ode} &
proposal correction and change-of-variables spine &
proposal density, target density, log determinant, corrected weight \\
Resampling and OT &
Chapter~\ref{ch:bf-dpf-resampling-ot}, Sections~\ref{sec:bf-dr-discontinuous-map}--\ref{sec:bf-dr-sinkhorn-numerics} &
\eqref{eq:bf-dr-two-particle-discontinuity}, \eqref{eq:bf-dr-soft-test-bias}, \eqref{eq:bf-dr-sinkhorn-marginals} &
differentiable resampling, OT, Sinkhorn sources &
ancestor path, soft surrogate, coupling residuals, barycentric cloud \\
Learned OT &
Chapter~\ref{ch:bf-dpf-learned-ot}, Sections~\ref{sec:bf-learned-ot-objects}--\ref{sec:bf-learned-ot-banking-evidence} &
\eqref{eq:bf-learned-ot-student-map}, \eqref{eq:bf-learned-ot-residual}, \eqref{eq:bf-learned-ot-equivariance} &
EOT teacher and set-architecture sources &
teacher output, student output, residuals, distribution tags \\
HMC target &
Chapter~\ref{ch:bf-dpf-hmc-target-suitability}, Sections~\ref{sec:bf-dpf-hmc-contract}--\ref{sec:bf-dpf-hmc-banking-gates} &
\eqref{eq:bf-dpf-hmc-contract} &
HMC, pseudo-marginal, surrogate-HMC spine &
scalar value, gradient, accept/reject value, target-status label \\
\bottomrule
\end{longtable}
\end{center}

\section{Promotion thresholds}
\label{sec:bf-dpf-debug-promotion-thresholds}

\begin{longtable}{@{}p{0.19\textwidth} p{0.30\textwidth} p{0.30\textwidth} p{0.13\textwidth}@{}}
\caption{Promotion thresholds for DPF implementation claims.}
\label{tab:bf-dpf-debug-promotion-thresholds}\\
\toprule
Claim level & Required passing evidence & Disallowed interpretation & Gate \\
\midrule
\endfirsthead
\toprule
Claim level & Required passing evidence & Disallowed interpretation & Gate \\
\midrule
\endhead
Exploratory implementation &
Layer-local smoke tests pass on minimal controlled examples; failures are
recorded with equation labels. &
Treating smoothness or runtime as target correctness. &
May run more experiments. \\
Credible research implementation &
Classical PF, flow, PF-PF, resampling/OT, learned-map, and HMC value-gradient
checks pass on named model classes with sensitivity reports. &
Generalizing beyond tested $N$, $d_x$, $\varepsilon$, model class, or target
status. &
May make bounded research claims. \\
Bank-facing research claim &
Research checks plus posterior comparisons, stress regimes, reproducible
artifacts, and documented failure policies. &
Claiming production suitability from a chapter or benchmark alone. &
Requires model-risk review. \\
Production or governance claim &
Independent review, change control, monitoring, rollback, data/model
governance, release stability, and ongoing diagnostics. &
Assuming academic derivation or local tests are sufficient approval. &
Requires separate governance process. \\
\bottomrule
\end{longtable}

\section{Boundary of this contract}
\label{sec:bf-dpf-debug-boundary}

This chapter does not validate any DPF implementation.  It defines the
minimum diagnostic evidence needed to interpret failures and to avoid promoting
a method beyond its evidence.  The word ``validate'' is therefore reserved for a
specific equation-local check or for a separately governed model-risk process;
it is not used here as a blanket claim about DPF, learned OT, HMC, or banking
deployment.

\FloatBarrier
