\chapter{Filter Choice}
\label{ch:bf-filter-choice}

Filter choice is a target-definition decision, not only a speed decision.  For
HMC, the selected backend determines the log likelihood, the derivative path,
the finite-failure policy, and the compilation surface.  BayesFilter should
choose the simplest backend that gives the required target fidelity and a
validated gradient.

\section{Decision Register}
\label{sec:bf-filter-choice-register}

The backend decision should be recorded in a short register:
\begin{description}
  \item[Exact Kalman, solve, or square-root LGSSM]
  Best for linear Gaussian likelihoods and regression oracles.  The main risks
  are SPD failures, mask handling, and large-scale factor failures.  The HMC
  gate is analytic/custom score parity plus compiled value-gradient parity.

  \item[EKF]
  Best as a cheap nonlinear baseline and affine recovery check.  The main risks
  are linearization bias and Jacobian fragility.  The HMC gate is affine Kalman
  recovery plus finite-gradient stress tests.

  \item[Sigma-point]
  Best for moderate nonlinear Gaussian approximations.  The main risks are an
  approximate target and covariance-update instability.  The HMC gate is
  reference recovery, static sigma-point shapes, and derivative parity.

  \item[Square-root sigma-point]
  Best for value-side robustness under near-singular covariance.  The main risk
  is false confidence when the implementation reconstructs covariance matrices
  internally.  The HMC gate is factor reconstruction, finite gradients, and
  compiled parity.

  \item[SVD sigma-point]
  Best for robust approximate value paths and spectral diagnostics.  The main
  risks are close or repeated spectral values and raw autodiff instability.  The
  HMC gate is eigen-gap telemetry, stress gates, and a custom-gradient fallback.

  \item[Particle or differentiable particle filter]
  Best for non-Gaussian diagnostics and frontier approximations.  The main
  risks are Monte Carlo variance, resampling non-differentiability, and relaxed
  target bias.  The HMC gate is an estimator-variance report and a source-backed
  correction policy.
\end{description}

\section{Recommended Order}
\label{sec:bf-filter-choice-order}

For a new model, BayesFilter should proceed in this order:
\begin{enumerate}[label=(\roman*)]
  \item reduce to an LGSSM or affine test case and validate the exact Kalman
    likelihood and derivatives;
  \item add masks, missing-data patterns, and mixed-frequency timing while
    retaining exact references;
  \item test EKF or sigma-point approximations on low-dimensional nonlinear
    examples with dense or trusted references;
  \item only then attempt DSGE-scale SVD sigma-point HMC;
  \item treat particle methods and transport methods as separate evidence
    streams unless a correction policy makes them target-preserving.
\end{enumerate}

\section{Industrial Default}
\label{sec:bf-filter-choice-industrial-default}

For NAWM-sized or industrial applications, the default should be an exact or
controlled approximate likelihood with an analytic or custom derivative path.
Raw tape gradients through spectral decompositions are acceptable as
diagnostic tools and small-case oracles, but they should not be the final
production plan unless spectral-gap, finite-gradient, and compiled parity
evidence survives the intended stress regime.

This policy is intentionally conservative.  The SVD sigma-point debugging work
showed that value robustness and HMC-gradient robustness are different
problems.  The next implementation phases should close that gap deliberately
instead of discovering it during long chains.
