\chapter{HMC for State-Space Likelihoods}
\label{ch:bf-hmc-for-state-space}

This chapter will develop Hamiltonian Monte Carlo and NUTS for filtering-based
posterior targets. The mathematical foundations are supported by the Phase 2B
literature gate, but implementation claims remain gated by BayesFilter's target,
gradient, and compilation contracts.

\section{Target Contract}
\label{sec:bf-hmc-target-contract}

For BayesFilter, an HMC transition is valid only relative to the target in
Definition~\ref{def:bf-filtering-target}.  The sampler needs the scalar log
target and its gradient in the coordinates used by the integrator.  If the
filter supplies a custom score, that score must correspond to the same log
likelihood value used in the Metropolis correction.

The HMC chapter therefore depends on the earlier derivative chapters.  A
filtering backend that can evaluate $\ell(\theta)$ but has no declared gradient
policy is value-only evidence.  It can be used for optimization, diagnostics,
or likelihood comparison, but it should not be promoted to HMC production use.

\section{Implementation Position}
\label{sec:bf-nuts-implementation-position}

TensorFlow Probability provides useful reference implementations of HMC and
NUTS, but BayesFilter should not treat TFP NUTS as the strategic production
backend. This is a project implementation decision, not a mathematical
criticism of NUTS. The decision is recorded in the source map and reset memo and
is supported by the minimal Gaussian benchmark in
\texttt{docs/benchmarks/benchmark\_tfp\_nuts\_gaussian.py}.

The reasons to keep TFP NUTS as a reference rather than the default production
path are:

\begin{itemize}
  \item full-trajectory NUTS control flow is hard to keep reliably XLA-compiled
    across model-specific likelihoods, adaptation logic, and diagnostics;
  \item nested TensorFlow Probability kernels can obscure whether the complete
    value-and-gradient pipeline is compiled or only partially traced;
  \item debugging numerical failures inside TFP kernel internals is slower than
    debugging a small BayesFilter-owned HMC/NUTS transition;
  \item large DSGE targets require explicit policies for finite log density,
    finite gradients, static shapes, filter failure handling, and custom
    derivative backends.
\end{itemize}

Consequently, this chapter will consolidate NUTS foundations from the approved
literature sources while treating TFP NUTS as a reference and diagnostic
baseline. BayesFilter's production direction is a small, inspectable HMC layer
whose contracts are designed around filtering likelihoods, analytic/custom
gradients, and XLA/JIT discipline. Any future NUTS implementation should satisfy
the same filter-aware contracts before it is promoted beyond diagnostic use.

\section{Diagnostic Use of NUTS}
\label{sec:bf-nuts-diagnostic-use}

NUTS remains valuable as an algorithmic reference and as a diagnostic sampler
on small controlled targets.  The bounded BayesFilter decision is narrower: the
project should not treat TFP NUTS as the default answer to every convergence,
geometry, or compilation problem.  The Gaussian benchmark shows that TFP NUTS
can XLA-compile on a toy target, but it is still materially heavier than
fixed-step HMC there.  Larger filtering targets add the hard parts omitted by
that benchmark: spectral derivative risk, missing-data branches, model-specific
invalid regions, and backend fallbacks.
