\chapter{Mass Matrices and Local Geometry}
\label{ch:bf-mass-matrices}

Mass matrices translate derivative information into sampler geometry.  In a
linear Gaussian validation problem, a local posterior Hessian or observed
information matrix can be a useful preconditioner.  In a nonlinear or weakly
identified model, the same matrix can be misleading unless it is regularized and
checked against diagnostics.

\section{Sources of Curvature}
\label{sec:bf-curvature-sources}

BayesFilter recognizes several curvature sources:
\begin{itemize}
  \item analytic likelihood Hessian plus prior curvature;
  \item observed-information or expected-information approximations;
  \item finite-difference Hessian diagnostics at a MAP point;
  \item empirical covariance from warmup draws;
  \item diagonal fallbacks when dense curvature is unstable.
\end{itemize}
The source must be recorded because each source has a different failure mode.
MacroFinance mass-matrix tests check eigenvalue regularization, dense versus
diagonal choices, whitening finiteness, and covariance construction from
Hessian helpers.  These are geometry utilities; they are not identification
proofs.

\section{Regularization Policy}
\label{sec:bf-mass-regularization}

A mass matrix used by HMC must be symmetric positive definite.  If a curvature
estimate is indefinite or ill-conditioned, BayesFilter should regularize it,
record the regularization, and expose the eigenvalue diagnostics.  Silently
using an indefinite curvature matrix as if it were a covariance is a target
geometry error.
