\chapter{XLA and JIT Compilation}
\label{ch:bf-xla-jit}

BayesFilter uses JIT compilation as a production tool, not as a correctness
argument.  A sampler is not fully compiled merely because an outer function is
wrapped in \texttt{tf.function}.  The complete value-and-gradient path must be
inside the compiled region, with static shapes and compatible device placement.

\section{Full-Pipeline Requirement}
\label{sec:bf-full-pipeline-jit}

For HMC, the compiled path includes:
\begin{itemize}
  \item parameter transform;
  \item model-provider maps;
  \item filtering likelihood;
  \item custom or autodiff gradient path;
  \item invalid-region branch;
  \item leapfrog transition or equivalent integrator.
\end{itemize}
If any component falls back to Python, retraces dynamically, or executes an
opaque noncompiled kernel, the implementation should be described as partially
compiled.

\section{Device and Shape Constraints}
\label{sec:bf-xla-device-shape}

The DSGE XLA source notes emphasize two lessons that carry over to
BayesFilter.  First, XLA compiles a function for a target platform, so mixing
CPU-only custom calls with GPU-targeted operations inside one JIT block can
fail.  Second, XLA requires static shapes.  Dynamic metadata should be encoded
as values or opaque compile-time metadata, not as changing tensor ranks.

The practical policy is phase separation: solve or model-construction phases
may be compiled separately from the filtering/HMC phase if they have different
device requirements.  A monolithic JIT block is valuable only when it actually
contains compatible operations.
