\chapter{Diagnostics and Failure Taxonomy}
\label{ch:bf-diagnostics}

Diagnostics are evidence about failures and geometry.  They are not proofs of
convergence.  BayesFilter should report diagnostics at three levels: filter
evaluation, derivative evaluation, and sampler behavior.

\section{Filter and Derivative Diagnostics}
\label{sec:bf-filter-derivative-diagnostics}

Filter diagnostics include factor pivots, reconstruction errors, solve
residuals, condition numbers, invalid-region counts, and finite-value checks.
Derivative diagnostics include score/Hessian finite checks, symmetry errors,
backend parity deltas, finite-difference residuals, and custom-gradient parity.
For SVD or eigendecomposition paths, singular-value or eigenvalue gap telemetry
is mandatory before HMC claims are made.

\section{Sampler Diagnostics}
\label{sec:bf-sampler-diagnostics}

Sampler diagnostics include acceptance rate, divergences, energy behavior,
effective sample size, split R-hat, mass-matrix conditioning, and tree depth if
NUTS is used diagnostically.  Short clean runs are smoke tests.  They can justify
the next experiment, but they do not establish convergence for an industrial
target.

\section{Failure Taxonomy}
\label{sec:bf-failure-taxonomy}

Common failure classes are:
\begin{itemize}
  \item target-domain failures;
  \item factorization or solve failures;
  \item non-finite value or gradient failures;
  \item spectral derivative instability;
  \item JIT retracing or partial compilation;
  \item sampler geometry failures;
  \item unsupported missing-data or mask patterns.
\end{itemize}
Each failure should point to a next action: fix the model contract, change the
backend, add jitter or regularization, strengthen derivative validation, or
downgrade the backend to diagnostic use.
