\chapter{Transport Maps and Surrogate Geometry}
\label{ch:bf-transport-surrogates}

Transport maps and surrogates are downstream geometry tools.  They can make an
HMC target easier to explore, but they cannot repair an invalid likelihood,
an inconsistent custom gradient, or a non-finite filter boundary.  This chapter
therefore depends on the target, derivative, and diagnostics contracts already
defined in the earlier chapters.

\section{Transported Targets}
\label{sec:bf-transported-targets}

Let $q\in\R^d$ denote the original unconstrained parameter and let
\[
  q = T_\psi(z)
\]
be an invertible differentiable map from transported coordinates $z$ to the
original coordinates.  If the original target density is
\[
  \pi_Q(q) \propto \exp[-U_Q(q)],
\]
then the transported target is
\begin{equation}
\label{eq:bf-transported-potential}
  U_Z(z)
  =
  U_Q(T_\psi(z))
  -
  \log\left|\det J_{T_\psi}(z)\right| .
\end{equation}
Running HMC in $z$-space is target-preserving only when the sampler evaluates
\eqref{eq:bf-transported-potential} and its gradient.  The Jacobian term is not
optional.  A learned map that is imperfect can still be useful as a
preconditioner, but omitting the change-of-variables correction changes the
target.

\section{NeuTra Position}
\label{sec:bf-neutra-position}

The Phase 2B literature gate accepted NeuTra only as transport and geometry
support.  That is exactly the right status for BayesFilter.  A NeuTra-like map
can improve posterior geometry by making the transformed target closer to a
simple reference distribution.  It is not a correctness guarantee, and it is
not a substitute for a valid filtering likelihood or a derivative provider
that matches the Metropolis value.

The strategic use is:
\begin{enumerate}[label=(\roman*)]
  \item certify the filtering target and gradient first;
  \item train or choose a transport map to improve geometry;
  \item run HMC on the Jacobian-corrected transported target;
  \item compare diagnostics and effective sample size per unit time to the
    untransported or affine baseline.
\end{enumerate}
If the map has poor tail behavior, bad inverse stability, or large Jacobian
curvature, it can make sampling worse even when it improves a training loss.

\section{Surrogate Targets}
\label{sec:bf-surrogate-targets}

A surrogate likelihood or surrogate Hamiltonian is a different object from a
transport map.  If the surrogate changes the scalar log target used in the
Metropolis correction, the Markov chain targets the surrogate posterior unless
a correction mechanism is added.  BayesFilter should record one of three
statuses for every surrogate:
\begin{description}
  \item[diagnostic surrogate] used only for exploration, initialization, or
    comparison; no posterior correctness claim is made;
  \item[corrected surrogate] used inside a delayed-acceptance or
    Metropolis-corrected scheme that evaluates the intended target before a
    sample is accepted;
  \item[target-defining surrogate] deliberately defines the approximate
    posterior being sampled, with approximation error documented separately.
\end{description}
This distinction prevents a common failure mode: using a fast approximate
filter or neural surrogate for gradients and then interpreting the resulting
chain as if it sampled the exact filtering posterior.

\section{Diagnostics}
\label{sec:bf-transport-diagnostics}

Transport diagnostics should be tied to sampler performance and target
integrity:
\begin{itemize}
  \item transformed-target finite value and finite gradient checks;
  \item log-Jacobian finite checks and inverse/forward reconstruction residuals;
  \item tail probes for large $\|z\|$;
  \item acceptance rate, divergences, E-BFMI, and effective sample size per
    second relative to an affine baseline;
  \item shell-level Hessian or force diagnostics where available.
\end{itemize}
These diagnostics explain why a map helps or fails.  They do not override a
failed target contract.

\section{BayesFilter Policy}
\label{sec:bf-transport-policy}

Transport and surrogate methods should enter BayesFilter after the exact or
controlled approximate likelihood path has been validated.  They are valuable
for large models because geometry can dominate HMC cost, but they sit on top of
the filter contract rather than underneath it.  In particular, they should not
be proposed as a fix for SVD spectral-gradient instability, missing derivative
guards, or partial XLA compilation.  Those problems must be solved at the
filter and sampler-contract layer first.
