\chapter{LGSSM Validation Ladder}
\label{ch:bf-lgssm-validation}

The linear Gaussian state-space model is the first validation target for
BayesFilter because it supplies an exact likelihood, controlled derivatives,
and tractable regression tests.  Every nonlinear or DSGE backend should be
reduced to an LGSSM special case before it is trusted in HMC.

\section{Evidence Already Available}
\label{sec:bf-lgssm-evidence}

The MacroFinance project provides the strongest local evidence stream:
\begin{itemize}
  \item exact prediction-error likelihood formulas and analytic derivative
    notes;
  \item covariance, solve, Cholesky square-root, QR square-root, TensorFlow, and
    SVD-flavored backends;
  \item finite-difference and autodiff parity tests on small models;
  \item large-scale backend-ladder tests that report reconstruction errors,
    solve residuals, QR pivots, and finite derivative deltas;
  \item missing-data policy tests that fail unsupported sparse panels before
    the likelihood silently changes meaning.
\end{itemize}
This evidence justifies using the LGSSM as BayesFilter's regression oracle.  It
does not yet replace the BayesFilter-specific packaging and public API tests
that will be needed when the code is moved into this project.

\section{Validation Rungs}
\label{sec:bf-lgssm-rungs}

The LGSSM ladder is:
\begin{enumerate}[label=(\roman*)]
  \item fixed-parameter likelihood agreement against an independent Kalman
    implementation;
  \item score agreement against finite differences and small-model autodiff;
  \item Hessian agreement where the backend claims observed-information
    support;
  \item backend parity among solve, Cholesky, QR, and TensorFlow paths;
  \item mask and mixed-frequency parity against dense special cases;
  \item compiled/eager parity for the full value-and-gradient target;
  \item short HMC plumbing checks with finite diagnostics;
  \item medium and strict HMC runs only after the preceding rungs pass.
\end{enumerate}

\section{Interpretation}
\label{sec:bf-lgssm-interpretation}

Passing the LGSSM ladder means the sampler interface, derivative convention,
and linear algebra backends are coherent on an exact target.  It does not prove
that a nonlinear approximation is accurate, and it does not prove that a DSGE
posterior has good HMC geometry.  It is the floor, not the ceiling.
