\chapter{NK SVD Case Study}
\label{ch:bf-nk-svd-case-study}

The NK SVD case study is a warning label with useful engineering evidence.  It
shows how far a numerically hardened SVD sigma-point path can get, and it also
shows why BayesFilter should not equate bounded HMC output with convergence or
industrial readiness.

\section{What Passed}
\label{sec:bf-nk-svd-passed}

The source reset memos record several meaningful passes:
\begin{itemize}
  \item focused SVD, target-boundary, adjoint, and generic nonlinear SSM
    contracts passed during the post-refactor work;
  \item generic SSM SVD likelihoods and gradients compile under production XLA;
  \item NK SVD fixed-solution and full-solve likelihood/gradient paths compile
    under XLA with explicit opt-in;
  \item tiny NK SVD production-XLA HMC smoke tests run to completion with finite
    output;
  \item generic nonlinear SSM medium HMC writes diagnostics with finite chains
    and zero divergences under tuned short-run defaults.
\end{itemize}
These results are valuable sampler-usability and compilation evidence.

\section{What Remains Blocked}
\label{sec:bf-nk-svd-blocked}

The same reset memos record why NK convergence remains blocked.  Earlier
bounded stress runs produced nonfinite adjoint and lam1 inputs, SVD conditioning
guard activations, nonfinite intermediate arrays, and extreme but finite
transformed-theta states.  The later production-XLA smoke gates do not erase
that artifact history, because they are tiny runs and not clean stress
experiments.

Therefore the next NK claim must be a clean stress gate, not a convergence
claim.  A valid clean gate requires no checked-nonfinite adjoint captures, no
nonfinite lam1 inputs, no nonfinite SVD-filter captures, no unexplained
transformed-theta scale explosion, and zero or separately justified divergences.

\section{Lesson for BayesFilter}
\label{sec:bf-nk-svd-lesson}

The case study separates four statuses:
\begin{description}
  \item[filter-correct] likelihood and gradients agree with controlled
    references;
  \item[sampler-usable] bounded HMC runs produce finite chains and useful
    diagnostics;
  \item[converged] strict multi-chain diagnostics pass on the intended target;
  \item[production-XLA-gated] the full value-and-gradient path compiles and runs
    under the declared production policy.
\end{description}
NK SVD has meaningful production-XLA and smoke evidence, but it is not yet a
converged case study.  That distinction should stay visible in every future
BayesFilter report.
