\chapter{CIP/AFNS Case Study}
\label{ch:bf-cip-afns-case-study}

The CIP/AFNS material is useful to BayesFilter because it is a large structured
linear Gaussian application with economically meaningful observation equations,
missing-data pressure, and analytic derivative requirements.  The economics
belongs in the source CIP monograph.  BayesFilter should import only the
filtering lessons.

\section{Filtering Lessons}
\label{sec:bf-cip-filtering-lessons}

The reusable lessons are:
\begin{itemize}
  \item affine or linearized macro-finance measurement systems can become large
    LGSSMs with structured observation matrices;
  \item yield-curve blocks create dense measurement panels with heterogeneous
    maturities and noise scales;
  \item missing observations must be represented by explicit masks, not by
    silent imputation or dynamic shape changes;
  \item analytic derivatives and backend parity matter because HMC mass
    matrices and MAP curvature depend on the same likelihood being sampled.
\end{itemize}

\section{BayesFilter Boundary}
\label{sec:bf-cip-boundary}

BayesFilter should not re-derive AFNS economics, currency-basis theory, or
asset-pricing restrictions in this monograph.  It should define the state-space
objects that an application supplies:
\[
  F_\theta,\quad Q_\theta,\quad H_\theta,\quad R_\theta,\quad
  m_{0|0,\theta},\quad P_{0|0,\theta},
\]
and then validate that the filter, derivative provider, masks, and HMC
diagnostics behave as promised.  This boundary keeps BayesFilter reusable and
prevents application monographs from becoming maintenance forks of the
filtering code.

\section{Readiness Use}
\label{sec:bf-cip-readiness-use}

CIP/AFNS should be used as a large LGSSM readiness case once the BayesFilter
package exists.  The required evidence is backend parity, missing-data policy,
finite analytic gradients and Hessians, MAP/mass-matrix diagnostics, and short
HMC plumbing checks.  It is not the right first target for SVD sigma-point HMC
claims, because its main reusable strength is exact or controlled linear
Gaussian filtering.
