\chapter{NAWM-Scale Design Target}
\label{ch:bf-nawm-design-target}

NAWM-scale models define the industrial target for BayesFilter.  They should
not be presented as completed evidence in this monograph pass.  Instead, they
set the design requirements for robust filtering and HMC on high-dimensional
state-space systems.

\section{Why NAWM Changes the Standard}
\label{sec:bf-nawm-standard}

As state dimension, parameter dimension, and observation dimension grow, several
risks become routine rather than exceptional:
\begin{itemize}
  \item covariance spectra become clustered or nearly singular;
  \item HMC proposals visit invalid or barely valid structural regions;
  \item raw autodiff through eigendecompositions or SVDs becomes fragile;
  \item compile boundaries and dynamic branches become performance bottlenecks;
  \item short smoke runs become less informative about long-chain behavior.
\end{itemize}
This is why BayesFilter should prefer analytic or custom derivative paths for
production filtering wherever feasible.

\section{Design Requirements}
\label{sec:bf-nawm-requirements}

A NAWM-grade BayesFilter backend needs:
\begin{enumerate}[label=(\roman*)]
  \item a fixed, documented target contract and parameter transform;
  \item explicit state-space shape contracts;
  \item finite invalid-region returns and finite gradients where HMC moves;
  \item analytic/custom gradients for exact or controlled approximate
    likelihoods;
  \item spectral-gap and factor diagnostics for any spectral backend;
  \item full-pipeline JIT evidence, not only outer-wrapper tracing;
  \item reproducible diagnostics and reset memos for every long experiment.
\end{enumerate}

\section{Hypothesis}
\label{sec:bf-nawm-hypothesis}

The working hypothesis is that NAWM-scale HMC will be more robust with
source-backed analytic or custom filtering gradients than with raw tape
gradients through spectral decompositions.  This is not yet a theorem or a
completed benchmark.  It is the prevention-first design target motivated by the
SVD debugging history and by the derivative risks documented in the Phase 2B
literature gate.
