\chapter{ResearchAssistant Workflows}
\label{app:bf-researchassistant-workflows}

ResearchAssistant is the literature-ingestion sidecar for this monograph.  It
should be used when a chapter needs primary-source support rather than local
implementation provenance.  It is not a replacement for reading the source and
checking whether the cited paper supports the exact BayesFilter claim.

\section{Workspace Policy}
\label{sec:bf-ra-workspace-policy}

Use a persistent, ignored workspace under the BayesFilter tree:
\[
  \texttt{/home/chakwong/BayesFilter/.research/ra-bayesfilter-monograph}.
\]
Downloaded papers, source packages, extracted text, and inbox files stay out of
Git.  Durable decisions belong in `docs/plans' result notes, the reset memo, and
the source map.

\section{Ingestion Workflow}
\label{sec:bf-ra-ingestion-workflow}

The standard workflow is:
\begin{enumerate}[label=(\roman*)]
  \item search or fetch the primary source;
  \item verify metadata against title, authors, venue, DOI or arXiv id;
  \item reject mismatches explicitly;
  \item extract the citation neighborhood or source labels relevant to the
    BayesFilter claim;
  \item write a short result note stating supported claims and limits;
  \item only then add or use a bibliography entry.
\end{enumerate}
The Phase 2B result note is the model: it accepted exact arXiv source packages
for HMC, NUTS, NeuTra, Matrix Backprop, Kitagawa score/Hessian material, and
the differentiable particle-filter frontier, while rejecting several metadata
mismatches.

\section{Claim Gate}
\label{sec:bf-ra-claim-gate}

A literature-heavy paragraph is draftable only when one of the following is
true:
\begin{itemize}
  \item an accepted ResearchAssistant note supports the claim and records its
    limits;
  \item a local source document plus code/test audit supports the claim;
  \item the paragraph is explicitly labeled as a project decision or an open
    hypothesis.
\end{itemize}
Raw `.bib' entries, downloaded PDFs, and search results are not claim support.

\section{Current Blockers}
\label{sec:bf-ra-current-blockers}

The current blockers are:
\begin{itemize}
  \item primary Julier--Uhlmann or equivalent UKF support for detailed
    sigma-point literature claims;
  \item primary PMCMC and pseudo-marginal support before particle-filter
    correctness claims;
  \item square-root filtering primary sources if final square-root
    bibliographic prose is expanded;
  \item a dedicated derivation/code audit before Kalman score and Hessian
    formulas are marked peer-review ready;
  \item industrial-scale SVD-HMC safety evidence before NAWM-scale claims.
\end{itemize}
Until these blockers are closed, the monograph should keep the conservative
wording used in the current draft.
